---
editor: 
  markdown: 
    wrap: 72
---

# Module 1: R Environment Setup and Spatial Data Import {#sec-module1}

This module establishes your R spatial epidemiology toolkit and teaches
you how to import, join, and inspect spatial and health data. By the end
of this module, you will have a clean `sf` object ready for mapping and
analysis.

**What you will learn**

- How to install and load the packages this tutorial relies on.
- What a *shapefile*, a *GeoPackage*, and a *raster* are, and how to
  import all three into R using `sf` and `terra`.
- Why the GeoPackage (`.gpkg`) format has largely replaced the legacy
  shapefile (`.shp`).
- How to preprocess (tidy) spatial data before analysis.
- How to join a table of health outcomes to spatial boundaries.
- How to inspect and manage the coordinate reference system (CRS).

A short note on *why* setup matters: most frustrating spatial errors —
maps that look blank, points plotted in the wrong ocean, joins that
silently drop countries — happen *before* any modelling. This module
builds good habits that will save you hours later.

```{r setup-m1}
#| include: false
library(sf)
library(terra)
library(tmap)
library(ggplot2)
library(dplyr)
library(mlspatial)
```

------------------------------------------------------------------------

## Install and Load R Packages {#sec-install}

**Purpose:** Install and attach the core spatial epidemiology toolkit.

A **package** is a collection of R functions and data bundled together.
Two steps are always involved:

1.  **Install** (once per machine): `install.packages()` downloads the
    package from a repository (CRAN by default) and stores it on your
    computer.
2.  **Load** (every session): `library()` attaches the package so its
    functions are available. You must re-load packages each time you
    start a new R session.

The table below shows the packages we will use throughout this tutorial
and their roles:

| Task | Package | Role |
|------------------------|------------------------|------------------------|
| Vector spatial data | `sf` | Simple features standard for polygons, points, lines |
| Raster spatial data | `terra` | Fast raster extraction and processing |
| Thematic mapping | `tmap`, `ggplot2` | Static and interactive maps |
| Data manipulation | `dplyr`, `tidyr`, `readr` | Tidy data wrangling |
| Spatial statistics | `spdep` | Neighbours, Moran's I, Getis-Ord Gi\* |
| Geographically weighted regression | `GWmodel` | Local regression modelling |
| Integrated ML + mapping | `mlspatial` | Import, map, model, evaluate |
| Machine learning engines | `randomForest`, `xgboost`, `e1071`, `caret` | RF, XGB, SVR algorithms |

```{r install-load}
#| eval: false
# Install any missing packages
pkgs <- c("sf", "terra", "tmap", "ggplot2", "dplyr", "tidyr", "readr",
          "spdep", "GWmodel", "mlspatial", "randomForest", "xgboost", 
          "e1071", "caret", "knitr", "rmarkdown")

install_if_missing <- function(p) {
  if (!require(p, character.only = TRUE, quietly = TRUE)) {
    install.packages(p, repos = "https://cloud.r-project.org/")
    library(p, character.only = TRUE)
  }
}
invisible(sapply(pkgs, install_if_missing))
```

::: callout-note
## How this code works

The helper function avoids re-installing packages you already have:

- `pkgs <- c(...)` creates a *character vector* of package names.
- `install_if_missing <- function(p)` defines a function taking one
  package name `p`.
- `require(p, character.only = TRUE, quietly = TRUE)` tries to load the
  package. It returns `TRUE` if the package is available and `FALSE` if
  not — importantly, it does **not** stop with an error like `library()`
  would.
- `!require(...)` therefore means "the package is *not* available". When
  that is true, we install it and then load it.
- `character.only = TRUE` tells `require()`/`library()` to treat `p` as
  a *value* (a string like `"sf"`) rather than as a literal name.
  Without it, R would look for a package literally called `p`.
- `sapply(pkgs, install_if_missing)` applies the function to every
  element of `pkgs` in turn (a one-line loop).
- `invisible()` hides the output of `sapply()`, keeping the console
  tidy.

You can run the same logic more manually for a single package:

``` r
install.packages("sf")
library(sf)
```
:::

::: callout-tip
## Speed Tip

If you are on a slow internet connection, install packages one at a time
and restart RStudio after each batch of 3–4 packages.
:::

**Expected output:** All packages load without error. `mlspatial` should
be version ≥ 0.1.0.

------------------------------------------------------------------------

## Load Administrative Boundary Data {#sec-load-shp}

**Purpose:** Import polygon shapefiles representing administrative
regions.

We demonstrate two approaches: (a) built-in data from `mlspatial`
(fastest for learning), and (b) `sf::st_read()` for your own shapefiles.
The second approach is expanded in the next section.

```{r load-boundaries}
# Approach A: Built-in Africa shapefile from mlspatial
data(africa_shp, package = "mlspatial")

# Inspect the object
class(africa_shp)
dim(africa_shp)

# Approach B: Import your own shapefile (commented out)
# ghana_regions <- sf::st_read("path/to/ghana_regions.shp")
```

::: callout-note
## Understanding the `sf` object

`africa_shp` is an `sf` (simple features) object. It is essentially a
normal **data frame** with one special extra column, `geometry`, that
stores the spatial coordinates of each feature. This is why `class()`
prints both `"sf"` and `"data.frame"`: you get all the familiar
data-frame tools (filtering, summarising, joining) *plus* spatial tools.

- Each **row** is one feature (here, one country).
- Each **column** other than `geometry` is an *attribute* (e.g. country
  name, population).
- `dim()` returns 54 rows (54 African countries) and 9 columns, one of
  which is `geometry`.
:::

**Key parameters of `st_read()`:** - `dsn` (data source name): the path
to the file or folder containing the data — a local file, a directory, a
zipped shapefile, or a URL. - `layer`: the layer name; needed whenever
the source holds more than one layer (a GeoPackage, a File Geodatabase,
or a folder of shapefiles). Supply it explicitly for `.gpkg` files. -
`quiet = TRUE` suppresses progress messages.

**Expected output:** `africa_shp` is an `sf` data frame with a
`geometry` column and attribute columns (including `NAME`).

------------------------------------------------------------------------

## Importing Your Own Spatial Data {#sec-import-own}

**Purpose:** Read an external shapefile, GeoPackage, or raster into R
and prepare (tidy) them for analysis. This is the skill you need once
you move from the bundled example data to your own project.

### Two fundamental data models {.unnumbered}

Spatial data comes in two flavours, and R uses a different package for
each:

| Model | What it stores | Package | Example |
|------------------|------------------|------------------|------------------|
| **Vector** | Discrete features as points, lines, or polygons (with attribute tables) | `sf` | Admin boundaries, health facilities, roads |
| **Raster** | A continuous grid of cells (pixels), each holding a value | `terra` | Elevation, temperature, land cover, rainfall |

A **shapefile** is a *vector* format. Despite the `.shp` extension, it
is actually several files that must travel together — `.shp` (geometry),
`.dbf` (attribute table), `.shx` (spatial index), and `.prj` (the CRS
definition). If you move only the `.shp` file, the data will not open.

A **GeoPackage** (`.gpkg`) is the modern container for the same kind of
vector data: a *single* file, built on SQLite, that can hold several
layers at once. It is covered in *Import a GeoPackage with `sf`* below.

A **raster** file (commonly `.tif`/`.tiff`, from "GeoTIFF") is a grid of
cells. Each cell covers a fixed area and holds a single value, which is
why rasters are ideal for continuous surfaces such as elevation or
climate.

### Import a shapefile with `sf` {#sec-import-vector}

We use a small shapefile that ships with the `terra` package (the 12
cantons of Luxembourg). Replace the path with your own `.shp` file if
you want to import your own data.

```{r import-vector}
# Get the path to a built-in example shapefile
shp_path <- system.file("ex/lux.shp", package = "terra")
shp_path

# Read it into R as an sf object
lux <- sf::st_read(shp_path, quiet = TRUE)

# Quick look
class(lux)
nrow(lux)
```

::: callout-note
## How `st_read()` works

`st_read()` is the single most important function for vector data. Give
it a file path and it returns an `sf` object, automatically reading the
geometry, the attribute table, and the CRS from the accompanying files.

- `dsn` is the path (here passed positionally as `shp_path`).
- `quiet = TRUE` silences the layer/feature progress report.
:::

### Import a GeoPackage with `sf` {#sec-import-gpkg}

A **GeoPackage** (`.gpkg`) is an open standard (OGC) container built on
SQLite. It stores one or more spatial layers in a *single* file, so
there are no sidecar files to lose. The `sf` package ships real
GeoPackages in its `gpkg/` folder, so there is nothing to download: we
use the 100 North Carolina counties in `nc.gpkg`. Conveniently, the
identical dataset also ships as a shapefile, so we can compare the two
formats on the same data.

```{r import-gpkg}
# Path to a GeoPackage that ships with sf (100 North Carolina counties)
gpkg_path <- system.file("gpkg/nc.gpkg", package = "sf")
basename(gpkg_path)

# Inspect the container before reading: layers, geometry, features, CRS
sf::st_layers(gpkg_path)

# Read the layer into R as an sf object
nc_gpkg <- sf::st_read(gpkg_path, quiet = TRUE)

# The same dataset also ships as a shapefile...
nc_shp <- sf::st_read(system.file("shape/nc.shp", package = "sf"), quiet = TRUE)

# ...so the two formats can be compared directly
class(nc_gpkg)
dim(nc_gpkg)
sf::st_crs(nc_gpkg)$epsg

# The geometry column keeps whatever name the file itself declares
names(nc_gpkg)[ncol(nc_gpkg)]   # "geom"
names(nc_shp)[ncol(nc_shp)]     # "geometry"

# The attribute tables are identical
identical(
  sf::st_drop_geometry(nc_gpkg)[order(nc_gpkg$FIPS), ],
  sf::st_drop_geometry(nc_shp)[order(nc_shp$FIPS), ]
)
```

Either way `st_read()` returns an `sf` object, so the choice of format
changes how the data is stored, not how you analyse it.

A GeoPackage can also hold layers with **no geometry at all** — it is a
SQLite database first and a spatial format second.

```{r import-gpkg-nospatial}
# A geometry-free layer comes back as a plain data frame
nosp <- sf::st_read(system.file("gpkg/nospatial.gpkg", package = "sf"), quiet = TRUE)
class(nosp)
```

::: callout-note
## How `st_layers()` and `st_read()` work with a GeoPackage

- `system.file("gpkg/nc.gpkg", package = "sf")` builds the path to a
  file inside the installed `sf` package. These examples live in
  `gpkg/`, not the more usual `extdata/`, so the subfolder must be named
  explicitly.
- `st_layers()` queries the container and reports every layer with its
  geometry type, feature count, field count, and CRS. Run it whenever
  you receive an unfamiliar `.gpkg` — the file name tells you nothing
  about how many layers are inside.
- `st_read(gpkg_path)` returns an `sf` object. When a GeoPackage holds
  more than one layer you must name the one you want with
  `layer = "..."`; otherwise `st_read()` silently returns the *first*
  layer, which may not be the one you intended.
- `st_drop_geometry()` strips the geometry and returns a plain data
  frame, which is why `identical()` can compare the two attribute tables
  directly. The `order()` calls align the rows by `FIPS` first, because
  the two files store them in a different order.
- The geometry column is named by the file rather than hard-coded: this
  GeoPackage layer calls it `geom`, whereas `st_read()` always names a
  shapefile's geometry column `geometry`. Both are ordinary `sf`
  objects.
- Nothing here is GeoPackage-specific: `st_read()` delegates to GDAL, so
  the same call reads `.gpkg`, `.shp`, `.geojson`, `.gdb`, and dozens of
  other formats. Only `dsn` changes.
:::

::: callout-note
## Why `.gpkg` is more modern than `.shp`

The ESRI Shapefile dates from the early 1990s and carries known design
limitations. A GeoPackage is a single, open, relational container that
avoids most of them:

| Feature | `.shp` (shapefile) | `.gpkg` (GeoPackage) |
|------------------------|------------------------|------------------------|
| Files on disk | A mandatory trio (`.shp`, `.shx`, `.dbf`) plus sidecars (`.prj`, `.cpg`, …) | One file |
| Layers per file | One | Many |
| Field name length | Truncated to 10 characters | Up to 255 characters |
| Field types | Limited; no true dates, no 64-bit integers | Full SQLite type system |
| Geometry types | One type per file | Mixable, multiple geometry columns |
| Missing values | No true `NULL`; blank numerics often arrive as `0` | `NULL` distinct from `0` |
| File size | 2 GB per component | No comparable practical ceiling |
| Text encoding | Ambiguous, needs a `.cpg` file | UTF-8 by definition |
| Keys and indexes | None | Primary keys, constraints, spatial indexes |
| Standard | De facto (ESRI) | Open (OGC / SQLite) |

Two of these bite spatial epidemiology directly. The **10-character
field name limit** silently truncates columns: none of the `nc` fields
are long enough to lose characters, but join keys and outcome names in
your own data may be, and a truncated key drops records without warning.
And with **no true `NULL`**, a genuinely missing incidence rate may be
stored and read back as `0`, deflating rates and pulling maps toward
zero. GeoPackage also keeps the CRS inside the file, so a stray `.prj`
cannot be lost in transit, and one file is far easier to version,
checksum, and share than five.

In practice: use `.gpkg` for anything you control, but expect to meet
`.shp` constantly in legacy data and in exports from older GIS software.
`st_read()` handles both, so the mental cost of switching is one file
path.
:::

::: callout-tip
## Where to find more GeoPackage examples

`sf` ships five more GeoPackages alongside `nc.gpkg`:

``` r
list.files(system.file("gpkg", package = "sf"))
```

They are handy for practice, and `b_pump.gpkg` is a small piece of
history — a single point in Soho, London (longitude -0.1367, latitude
51.5133) marking the Broad Street water pump whose removal ended John
Snow's 1854 cholera outbreak and helped establish epidemiology as a
discipline.

When a data portal offers a choice, prefer `.gpkg`, then `.geojson`,
then a zipped `.shp`. GeoJSON is also a single file and universally
supported, but it is plain text, so it is much larger and slower than
`.gpkg` for polygon layers. To convert a shapefile you already have,
`sf::st_write(myshape, "out.gpkg")` is all it takes — no external GIS
software required.
:::

### Inspect and tidy the vector data {#sec-tidy-vector}

Before analysis, check the object's geometry type, CRS, and validity,
and tidy up anything that will cause trouble later.

```{r tidy-vector}
# Geometry type and number of features
sf::st_geometry_type(lux, by_geometry = FALSE)

# Coordinate reference system (EPSG code)
sf::st_crs(lux)

# Are all geometries valid? (FALSE means some need fixing)
all(sf::st_is_valid(lux))

# Fix any invalid geometries
lux <- sf::st_make_valid(lux)

# Drop empty geometries (features with no coordinates)
lux <- lux[!sf::st_is_empty(lux), ]

# Tidy column names: lower-case, underscores instead of spaces
names(lux) <- gsub("[[:space:]]+", "_", tolower(names(lux)))
names(lux)

# Keep only the columns you need (canton name and population)
lux_sub <- lux[, c("name_2", "pop", "geometry")]
head(lux_sub)
```

::: callout-note
## What each step does

- `st_geometry_type(..., by_geometry = FALSE)` reports the *single*
  shared geometry type (e.g. `MULTIPOLYGON`) rather than one per row.
- `st_crs()` shows the coordinate reference system (see the CRS section
  below).
- `st_is_valid()` checks each geometry for self-intersections and other
  defects; `st_make_valid()` repairs them. Invalid geometries are a
  common source of cryptic errors later.
- `st_is_empty()` flags features with no coordinates; we remove them.
- `gsub("[[:space:]]+", "_", tolower(names(lux)))` converts names like
  `"NAME_2"` to `"name_2"` and replaces spaces with underscores, so they
  are easier to type and safe to use in formulas.
- Subsetting with `lux[, c("name_2", "pop", "geometry")]` keeps only
  relevant columns (note `geometry` stays so it remains spatial).
:::

### Import a raster with `terra` {#sec-import-raster}

We use a small digital elevation model (DEM) that ships with `terra`.
Elevation is a common covariate in spatial epidemiology — for example,
altitude influences temperature and the distribution of vector-borne
diseases.

```{r import-raster}
# Path to a built-in example raster (a DEM)
r_path <- system.file("ex/elev.tif", package = "terra")
r_path

# Read it into R as a SpatRaster
elev <- terra::rast(r_path)

# Inspect it
elev
terra::nlyr(elev)     # number of layers (bands)
terra::res(elev)      # cell size in the x and y directions
terra::ext(elev)      # spatial extent (bounding box)
terra::crs(elev, describe = TRUE)$code  # EPSG code

# Plot it
plot(elev, main = "Elevation (m)")
```

::: callout-note
## Understanding the raster print-out

`terra::rast()` returns a `SpatRaster`. When you print it, you see:

- **dimensions** — number of rows, columns, and layers. This example has
  one layer (a single-band DEM); a multispectral satellite image might
  have several.
- **resolution** — the ground size of one cell (here ≈ 0.0083 degrees).
- **extent** — the bounding box (xmin, xmax, ymin, ymax).
- **crs** — the coordinate reference system.
:::

### Preprocess the raster {#sec-tidy-raster}

Rasters from different sources rarely line up. The three most common
preprocessing tasks are **reprojecting**, **cropping/masking**, and
**resampling**.

```{r tidy-raster}
# Reproject the raster and polygons to a shared projected CRS (EPSG:3857)
elev_web <- terra::project(elev, "EPSG:3857")
lux_web <- sf::st_transform(lux_sub, crs = 3857)

# Convert the sf polygons to a SpatVector so terra can use them
lux_vect <- terra::vect(lux_web)

# Crop to the extent of the polygons, then mask to their exact shape
elev_lux <- terra::crop(elev_web, lux_vect)
elev_lux <- terra::mask(elev_lux, lux_vect)

plot(elev_lux, main = "Elevation clipped to Luxembourg")
```

::: callout-note
## Why these three steps matter

- `project()` changes the CRS of a raster; `st_transform()` does the
  same for vector data. Layers must share a CRS before you can overlay
  or stack them.
- `crop()` trims the raster to a bounding box (fast); `mask()` then sets
  cells *outside* the polygon to `NA`, giving the raster the exact shape
  of your study area.
- `vect()` converts an `sf` object into a `terra`-compatible
  `SpatVector` (and `sf::st_as_sf()` converts back the other way).

The `terra` package also provides `resample()` to align rasters that
share a CRS but differ in resolution, which you will need when combining
multiple raster layers.
:::

::: callout-tip
## The `mlspatial` shortcuts

`mlspatial` ships thin wrappers over these steps for convenience:
`load_shapefile("path.shp")` (a wrapper around `st_read()`) and
`load_incidence_data("path.xlsx")`. Because `load_shapefile()` passes
the path straight through to `st_read()`, it reads a GeoPackage just as
happily — `load_shapefile("path.gpkg")` returns the first layer. Under
the hood these wrappers do exactly what we did above, so understanding
the `sf`/`terra` functions first is important.
:::

------------------------------------------------------------------------

## Load Health / Disease Incidence Data {#sec-load-health}

**Purpose:** Import tabular disease data and link it to spatial
boundaries.

```{r load-health}
# Built-in pancreatic cancer incidence data from mlspatial
data(panc_incidence, package = "mlspatial")

# Inspect structure
str(panc_incidence)
head(panc_incidence)
```

::: callout-note
## Reading `str()` and `head()`

- `str()` (structure) lists each column, its type, and the first few
  values. Use it to confirm that columns you expect to be numeric really
  are numeric, and to spot factor/character mix-ups.
- `head()` shows the first six rows so you can sanity-check the values.
:::

**Key variables in `panc_incidence`:** - `NAME`: Country name (the join
key) - `incidence`: Incidence rate per 100,000 population (as documented
in `mlspatial`) - `female`, `male`: Counts by sex - `agea`, `ageb`,
`agec`: Age-stratum values (below 20, 20–54, and above 55 years) - `yra`
… `yre`: Yearly incidence rates (2017–2021)

**Expected output:** A tibble with one row per country (54 rows) and 18
columns.

------------------------------------------------------------------------

## Join Health Data to Spatial Data {#sec-join}

**Purpose:** Attach disease attributes to the correct spatial polygons
so we can map and model them.

A join matches rows from two tables using a common **key** column. Here
the key is the country name. Crucially, the spelling must match exactly
— in our data the shapefile spells the country "Côte d'Ivoire" (with an
accent), while the health table spells it "Cote d'Ivoire" (without).
Without fixing this, the join would silently drop the country. We fix
the name first.

```{r join-data}
# Fix the one mismatched name so the join keys align
africa_shp <- africa_shp |>
  mutate(
    NAME       = gsub("Côte d'Ivoire", "Cote d'Ivoire", NAME),
    COUNTRYAFF = gsub("Côte d'Ivoire", "Cote d'Ivoire", COUNTRYAFF)
  )

# Method 1: mlspatial wrapper (recommended for consistency)
africa_health <- mlspatial::join_data(
  sf_data = africa_shp,
  tbl_data = panc_incidence,
  by = "NAME"
)

# Method 2: dplyr inner_join (identical result, more familiar syntax)
# africa_health <- africa_shp |>
#   dplyr::inner_join(panc_incidence, by = "NAME")

# Verify the join
dim(africa_health)
```

::: callout-note
## How the code works

- `gsub(pattern, replacement, x)` performs a **g**lobal
  **sub**stitution: it replaces every occurrence of `"Côte d'Ivoire"`
  with `"Cote d'Ivoire"` in the `NAME` column. We do the same for
  `COUNTRYAFF` so every name column is consistent.
- `mutate()` (from `dplyr`) creates or overwrites columns.
- `join_data()` is an `mlspatial` convenience wrapper; inspecting its
  source shows it simply calls `dplyr::inner_join()`.
- `by = "NAME"` tells R which column is the shared key. Both tables
  **must** use identical spelling and casing.
- `inner_join()` keeps only rows that match in both tables. Use
  `left_join()` if you want to keep all polygons even those without
  health data (unmatched rows get `NA`).
:::

**Expected output:** An `sf` object with the same number of rows as the
original shapefile (54, because every country matched) and all columns
from both datasets.

------------------------------------------------------------------------

## Explore the Joined Dataset {#sec-explore}

**Purpose:** Confirm that the join worked and understand the data
structure.

```{r explore-joined}
# Structure of the sf object
str(africa_health)

# First six rows
head(africa_health, 6)

# Summary of the incidence variable
summary(africa_health$incidence)

# Check for missing values in key variables
sum(is.na(africa_health$incidence))
```

::: callout-note
## Reading the output

- `str()` should show a `Classes 'sf' ...` object with a `geometry`
  column of type `MULTIPOLYGON`.
- `summary()` gives the five-number summary plus the mean for a numeric
  column. `incidence` is numeric and ranges from about 1.8 to 4950.9,
  with a median near 185.
- `sum(is.na(...))` counts missing values. `0` means every country has a
  recorded incidence, so no imputation is needed here.
:::

**Expected output:** - `str()` shows a `Classes 'sf' ...` object with a
`geometry` column of type `MULTIPOLYGON`. - `incidence` is numeric with
a range of roughly 1.8 to 4900 and no missing values.

------------------------------------------------------------------------

## Coordinate Reference System (CRS) Management {#sec-crs}

**Purpose:** Ensure all spatial layers share the same projection.
Mismatched CRS is the most common source of silent errors in spatial
analysis.

A **coordinate reference system** defines how the flat coordinates in
your data relate to locations on the Earth. It has two parts: a
**datum** (a model of the Earth's shape) and a **projection** (how the
curved Earth is flattened onto a map). CRSs are identified by **EPSG
codes** (e.g. `4326`, `3857`).

```{r crs-management}
# View current CRS
sf::st_crs(africa_health)

# Check if it is geographic (degrees) or projected (meters)
sf::st_is_longlat(africa_health)

# If you need to transform to a projected CRS for distance calculations
# Example: Africa Albers Equal Area Conic (EPSG: 102022)
# africa_projected <- sf::st_transform(africa_health, crs = 102022)
```

::: callout-note
## Geographic vs projected CRS

- **Geographic** CRSs (e.g. EPSG:4326, WGS 84) use degrees of
  latitude/longitude. `st_is_longlat()` returns `TRUE` for these.
  Degrees are *not* a consistent unit of distance (a degree of longitude
  shrinks toward the poles), so they are unsuitable for measuring
  distances or areas.
- **Projected** CRSs (e.g. EPSG:3857, or UTM zones) use metres on a flat
  surface. Distances and areas are meaningful, but every projection
  distorts *something* (shape, area, or distance).
- `st_transform()` converts between CRSs. It changes the stored
  coordinates but not the underlying data.
:::

**Key concepts:** - **EPSG:3857** (WGS 84 / Pseudo-Mercator): a web
mapping projection in metres, used by Google Maps and OpenStreetMap. It
is convenient for visualisation but distorts areas and distances, which
worsens away from the equator. - **Projected CRS** (e.g. UTM zones,
Albers Equal Area): distances in metres, with distortion controlled for
a specific region. Essential for buffers, distances, and GWR bandwidth
selection.

For this tutorial we keep EPSG:3857 because `tmap` handles it well for
visualisation, but we note that distance calculations may be distorted.

**Expected output:** `st_crs()` returns the CRS definition (EPSG:3857,
WGS 84 / Pseudo-Mercator). `st_is_longlat()` returns `FALSE` because
this is a projected CRS with units in metres, not geographic degrees.

------------------------------------------------------------------------

## Practical Exercise {#sec-m1-exercise}

Now apply what you have learned. Work through the tasks below, then
check your answers against the solution.

1.  Load `africa_shp` and report its class, dimensions, geometry type,
    and CRS code.
2.  Import the bundled Luxembourg shapefile
    (`system.file("ex/lux.shp", package = "terra")`). How many cantons
    does it contain, and what geometry type does it use?
3.  Import the bundled elevation raster
    (`system.file("ex/elev.tif", package = "terra")`) and report its
    resolution, extent, and CRS code.
4.  Re-project that raster (in Q3) to EPSG:3857 and confirm the new CRS.
5.  Recreate the country join yourself using `inner_join()` (without
    `mlspatial::join_data()`), remembering to fix the "Côte d'Ivoire"
    spelling first. How many rows does the result have?
6.  Read the GeoPackage that ships with `sf`
    (`system.file("gpkg/nc.gpkg", package = "sf")`) and use
    `st_layers()` to report its layer name, feature count, and geometry
    type. What is that layer's geometry column called? Then read the
    equivalent shapefile (`system.file("shape/nc.shp", package = "sf")`)
    and explain why both objects behave identically.
7.  *(Challenge)* Use `st_is_valid()` on `africa_shp` and report how
    many countries (if any) have invalid geometries. If any are invalid,
    repair them with `st_make_valid()`.

::: {.callout-note collapse="true"}
## Model solution

``` r
library(sf)
library(terra)
library(dplyr)
library(mlspatial)

# 1. Built-in Africa shapefile
data(africa_shp, package = "mlspatial")
class(africa_shp)                       # "sf" "data.frame"
dim(africa_shp)                         # 54 rows x 9 columns
sf::st_geometry_type(africa_shp, by_geometry = FALSE)  # MULTIPOLYGON
sf::st_crs(africa_shp)$epsg            # 3857

# 2. Bundled Luxembourg shapefile
lux <- sf::st_read(system.file("ex/lux.shp", package = "terra"), quiet = TRUE)
nrow(lux)                               # 12 cantons
sf::st_geometry_type(lux, by_geometry = FALSE)  # POLYGON

# 3. Bundled elevation raster
elev <- terra::rast(system.file("ex/elev.tif", package = "terra"))
terra::res(elev)                        # 0.008333333 0.008333333
terra::ext(elev)                        # 5.741667 6.533333 49.44167 50.19167
terra::crs(elev, describe = TRUE)$code  # 4326

# 4. Reproject
elev_web <- terra::project(elev, "EPSG:3857")
terra::crs(elev_web, describe = TRUE)$code  # 3857

# 5. Manual join
data(panc_incidence, package = "mlspatial")
africa_fixed <- africa_shp |>
  mutate(NAME = gsub("Côte d'Ivoire", "Cote d'Ivoire", NAME))
africa_health2 <- dplyr::inner_join(africa_fixed, panc_incidence, by = "NAME")
nrow(africa_health2)                    # 54

# 6. GeoPackage shipped with sf
gpkg_path <- system.file("gpkg/nc.gpkg", package = "sf")
sf::st_layers(gpkg_path)                # layer "nc.gpkg", 100 features, MULTIPOLYGON
nc_gpkg <- sf::st_read(gpkg_path, quiet = TRUE)
names(nc_gpkg)[ncol(nc_gpkg)]           # "geom"

nc_shp <- sf::st_read(system.file("shape/nc.shp", package = "sf"), quiet = TRUE)
nrow(nc_shp)                            # 100 - same data, geometry column "geometry"

# 7. Validity check
sum(!sf::st_is_valid(africa_shp))       # number of invalid geometries
africa_shp <- sf::st_make_valid(africa_shp)
```
:::

------------------------------------------------------------------------

## Module 1 Checklist {#sec-m1-check .unnumbered}

- [ ] All packages installed and loaded
- [ ] `africa_shp` imported successfully
- [ ] You can import an external shapefile with `st_read()`
- [ ] You can read a GeoPackage and explain why `.gpkg` supersedes
  `.shp`
- [ ] You can import and preprocess an external raster with `terra`
- [ ] `panc_incidence` loaded and inspected
- [ ] `africa_health` created via `join_data()`
- [ ] CRS verified and understood
- [ ] Practical exercise completed

**Next:** Proceed to [Module 2: Disease Mapping in R](02-mapping.qmd).
