---
editor: 
  markdown: 
    wrap: 72
---

# R Environment Setup and Spatial Data Import {#sec-module1}

This module establishes your R spatial epidemiology toolkit and teaches
you how to import, join, and inspect spatial and health data. By the end
of this module, you will have a clean `sf` object ready for mapping and
analysis.

```{r setup-m1}
#| include: false
library(sf)
library(terra)
library(tmap)
library(ggplot2)
library(dplyr)
library(mlspatial)
```

------------------------------------------------------------------------

## Install and Load R Packages {#sec-install}

**Purpose:** Install and attach the core spatial epidemiology toolkit.

The table below shows the packages we will use throughout this tutorial
and their roles:

| Task | Package | Role |
|------------------------|------------------------|------------------------|
| Vector spatial data | `sf` | Simple features standard for polygons, points, lines |
| Raster spatial data | `terra` | Fast raster extraction and processing |
| Thematic mapping | `tmap`, `ggplot2` | Static and interactive maps |
| Data manipulation | `dplyr`, `tidyr`, `readr` | Tidy data wrangling |
| Spatial statistics | `spdep` | Neighbours, Moran's I, Getis-Ord Gi\* |
| Geographically weighted regression | `GWmodel` | Local regression modelling |
| Integrated ML + mapping | `mlspatial` | Import, map, model, evaluate |
| Machine learning engines | `randomForest`, `xgboost`, `e1071`, `caret` | RF, XGB, SVR algorithms |

```{r install-load}
#| eval: false
# Install any missing packages
pkgs <- c("sf", "terra", "tmap", "ggplot2", "dplyr", "tidyr", "readr",
          "spdep", "GWmodel", "mlspatial", "randomForest", "xgboost", 
          "e1071", "caret", "knitr", "rmarkdown")

install_if_missing <- function(p) {
  if (!require(p, character.only = TRUE, quietly = TRUE)) {
    install.packages(p, repos = "https://cloud.r-project.org/")
    library(p, character.only = TRUE)
  }
}
invisible(sapply(pkgs, install_if_missing))
```

::: callout-tip
## Speed Tip

If you are on a slow internet connection, install packages one at a time
and restart RStudio after each batch of 3–4 packages.
:::

**Expected output:** All packages load without error. `mlspatial` should
be version ≥ 0.1.0.

------------------------------------------------------------------------

## Load Administrative Boundary Data {#sec-load-shp}

**Purpose:** Import polygon shapefiles representing administrative
regions.

We demonstrate two approaches: (a) built-in data from `mlspatial`
(fastest for learning), and (b) `sf::st_read()` for your own shapefiles.

```{r load-boundaries}
# Approach A: Built-in Africa shapefile from mlspatial
data(africa_shp, package = "mlspatial")

# Inspect the object
class(africa_shp)
dim(africa_shp)

# Approach B: Import your own shapefile (commented out)
# ghana_regions <- sf::st_read("path/to/ghana_regions.shp")
```

**Key parameters:** - `st_read()` arguments: `dsn` (data source name /
folder path), `layer` (filename without .shp extension if inside a
geodatabase). - `quiet = TRUE` suppresses progress messages.

**Expected output:** `africa_shp` is an `sf` data frame with a
`geometry` column and attribute columns (including `NAME`).

------------------------------------------------------------------------

## Load Health / Disease Incidence Data {#sec-load-health}

**Purpose:** Import tabular disease data and link it to spatial
boundaries.

```{r load-health}
# Built-in pancreatic cancer incidence data from mlspatial
data(panc_incidence, package = "mlspatial")

# Inspect structure
str(panc_incidence)
head(panc_incidence)
```

**Key variables in `panc_incidence`:** - `NAME`: Country name (join
key) - `incidence`: Age-standardized rate per 100,000 - `female`,
`male`: Counts by sex - `agea`, `ageb`, `agec`: Age-stratum counts -
`yra` … `yre`: Yearly incidence rates (2017–2021)

**Expected output:** A tibble with one row per country and columns for
incidence, demographics, and age groups.

------------------------------------------------------------------------

## Join Health Data to Spatial Data {#sec-join}

**Purpose:** Attach disease attributes to the correct spatial polygons
so we can map and model them.

```{r join-data}
# Method 1: mlspatial wrapper (recommended for consistency)
africa_health <- mlspatial::join_data(
  sf_data = africa_shp |> 
    # change ivory coast name to match in both datasets
    mutate(
      across(c(NAME, COUNTRYAFF),
             ~ replace_values(.x, "Côte d'Ivoire" ~ "Cote d'Ivoire")
     )
    ),
  tbl_data = panc_incidence,
  by = "NAME"
)

# Method 2: dplyr inner_join (identical result, more familiar syntax)
# africa_health <- africa_shp |> 
#   dplyr::inner_join(panc_incidence, by = "NAME")

# Verify the join
dim(africa_health)
```

**Key parameters:** - `by`: Character name of the shared key column.
Both tables **must** share identical spelling and casing. -
`inner_join()` keeps only rows that match in both tables. Use
`left_join()` if you want to keep all polygons even without health data.

**Expected output:** An `sf` object with the same number of rows as the
original shapefile (or fewer if using `inner_join` and some names do not
match) and all columns from both datasets.

------------------------------------------------------------------------

## Explore the Joined Dataset {#sec-explore}

**Purpose:** Confirm that the join worked and understand the data
structure.

```{r explore-joined}
# Structure of the sf object
str(africa_health)

# First six rows
head(africa_health, 6)

# Summary of the incidence variable
summary(africa_health$incidence)

# Check for missing values in key variables
sum(is.na(africa_health$incidence))
```

**Expected output:** - `str()` shows a `Classes 'sf' ...` object with a
`geometry` column of type `MULTIPOLYGON`. - `incidence` is numeric with
a range roughly between 0.5 and 8.0 (rates per 100,000).

------------------------------------------------------------------------

## Coordinate Reference System (CRS) Management {#sec-crs}

**Purpose:** Ensure all spatial layers share the same projection.
Mismatched CRS is the most common source of silent errors in spatial
analysis.

```{r crs-management}
# View current CRS
sf::st_crs(africa_health)

# Check if it is geographic (degrees) or projected (meters)
sf::st_is_longlat(africa_health)

# If you need to transform to a projected CRS for distance calculations
# Example: Africa Albers Equal Area Conic (EPSG: 102022)
# africa_projected <- sf::st_transform(africa_health, crs = 102022)

# For this tutorial we keep WGS84/Pseudo-Mercator (EPSG:3857) because tmap 
# handle it well for visualisation, but we note that distance calculations may
# be distorted due to projection limitations.
```

**Key concepts:** - **EPSG:3857** (WGS84/Pseudo-Mercator): Web mapping projection in meters. Good for visualisation and web mapping, but distances are distorted especially away from the equator. Not recommended for accurate distance calculations - **Projected
CRS** (e.g., UTM zones, Albers, Pseudo-Mercator): Distances in meters. Essential for
buffers, distances, and GWR bandwidth selection. However Pseudo-Mercartor distorts areas and distances significantly.

**Expected output:** `st_crs()` returns the CRS definition, shows EPSG:3857 (WGS 84/Pseudo-Mercator).`st_is_longlat()` returns `FALSE` because this is a projected CRS (Units in meters), not geographic degrees.

------------------------------------------------------------------------

## Module 1 Checklist {#sec-m1-check .unnumbered}

- [ ] All packages installed and loaded
- [ ] `africa_shp` imported successfully
- [ ] `panc_incidence` loaded and inspected
- [ ] `africa_health` created via `join_data()`
- [ ] CRS verified and understood

**Next:** Proceed to [Module 2: Disease Mapping in R](02-mapping.qmd).
