---
editor: 
  markdown: 
    wrap: sentence
---

# Module 2: Disease Mapping in R {#sec-module2}

This module teaches you to create publication-quality thematic maps using `mlspatial`, `tmap`, and `ggplot2`.
You will learn static maps, interactive maps, multi-layer overlays, and export techniques.

```{r setup-m2}
#| include: false
library(sf)
library(terra)
library(tmap)
library(ggplot2)
library(dplyr)
library(mlspatial)

data(africa_shp, package = "mlspatial")
data(panc_incidence, package = "mlspatial")
africa_health <- mlspatial::join_data(africa_shp |> 
    mutate(across(c(NAME, COUNTRYAFF),
                  ~ replace_values(.x, "Côte d'Ivoire" ~ "Cote d'Ivoire"))
    ), panc_incidence, by = "NAME")
```

------------------------------------------------------------------------

## Quick Thematic Maps with `mlspatial` {#sec-quick-map}

**Purpose:** Generate a publication-ready choropleth map in one line.

```{r quick-map-mlspatial}
map1 <- mlspatial::plot_single_map(
  sf_data = africa_health,
  var = "incidence",
  title = "Incidence (per 100,000)",
  palette = "yl_or_rd"
)
map1
```

**Key parameters:** - `var`: Column name to map (must be quoted).
- `palette`: Any RColourBrewer or viridis palette name.
`"yl_or_rd"`( old name = `"YlOrRd"`) is standard for public health "hot" maps.

**Expected output:** A static tmap object showing African countries coloured by pancreatic cancer incidence.

------------------------------------------------------------------------

## Add Country Labels to Maps {#sec-labels}

**Purpose:** Annotate maps so stakeholders can identify regions without a separate legend.

```{r labels-qtm}
labelled_map <- tm_shape(africa_health) +
  # Fill layer (colours the polygons)
  tm_fill(
    "incidence", 
    fill.scale = tm_scale(values = "brewer.yl_or_rd"),  
    fill.legend = tm_legend(title = "Rate per 100k")    
  ) +
  # Border layer (outlines)
  tm_borders(col = "grey50", lwd = 0.6) +
  # Text layer (adds, country names)
  tm_text(
    "NAME", 
    size = 0.6, 
    col = "black"
  ) +
  #  Title layer
  tm_title("Pancreatic Cancer Incidence in Africa",
           color = 'black', size = 1.2) + 
  # Layout (everything else about map appearance)
  tm_layout(
    legend.position = c("left", "bottom")
  ) +
  # Add Compass symbol
  tm_compass(
    type = "8star", 
    position = c("right", "bottom")
  ) +
  # Add scalebar 
  tm_scalebar(
    position = c("right", "bottom")
  )
labelled_map


# You can also create a quick map using the qtm() function (less flexible)
# qtm(africa_health, fill = "incidence", fill.scale = tm_scale(values = "brewer.yl_or_rd"))

```

**Key parameters:** - `tm_text()`: `text = "NAME"` specifies the label column.
- `size` is in map units; adjust downward for small countries.

**Expected output:** A choropleth map with country names overlaid at centroids.

------------------------------------------------------------------------

## Create Disease Distribution Choropleth Maps with `ggplot2` {#sec-ggplot}

**Purpose:** Leverage the tidyverse grammar of graphics for highly customized static maps.

```{r ggplot-choropleth}
library(ggplot2)

ggplot(africa_health) +
  geom_sf(aes(fill = incidence), colour = "white", linewidth = 0.2) +
  scale_fill_distiller(
    palette = "YlOrRd", direction = 1,
    name = "Incidence\n(per 100,000)",
    breaks = c(0, 2000, 4000)
  ) +
  theme_minimal() +
  labs(
    title = "Pancreatic Cancer Incidence in Africa",
    subtitle = "Age-standardized rate, 2021 estimates",
    caption = "Source: GBD 2021 via mlspatial",
    x = "", y = ""
  ) +
  # geom_text(
  #   aes(label = stringr::str_wrap(NAME, 8), geometry = geometry),
  #   stat = "sf_coordinates",
  #   colour = 'black', size = 3
  # ) +
  geom_sf_text(
    aes(label = stringr::str_wrap(NAME, 8)),
    size = 3.2, colour = 'black', 
    check_overlap = TRUE  # avoids overlapping labels
  ) +
  theme(
    legend.position = "bottom",
    plot.title = element_text(face = "bold", size = 14)
  ) +
  guides(
    fill = guide_colourbar(barheight = 0.5, barwidth = 8)
  )
```

**Key parameters:** - `geom_sf()`: The workhorse for sf objects.
`aes(fill = incidence)` maps the variable to color.
- `scale_fill_distiller()`: For continuous data.
Perceptually uniform, color-blind safe, and prints well in greyscale.

**Expected output:** A static ggplot map with a continuous colour legend.

------------------------------------------------------------------------

## Create Interactive Maps with `tmap` {#sec-interactive}

**Purpose:** Allow panning, zooming, and layer toggling for exploratory data analysis.

```{r interactive-tmap}
#| eval: false
# Switch tmap to interactive mode
tmap_mode("view")

interactive_map <- tm_shape(africa_health) +
  tm_fill(
    "incidence", 
    fill.scale = tm_scale(values = "brewer.yl_or_rd"), 
    fill.legend = tm_legend(title = "Incidence (per 100k)"),
    popup.vars = c("incidence", "female", "male")
  ) +
  tm_borders(col = "black", lwd = 0.5)

interactive_map

# Return to static mode for publication figures
tmap_mode("plot")
```

**Key parameters:** - `tmap_mode("view")`: Renders to Leaflet JavaScript widget.
- `popup.vars`: Vector of column names shown when a user clicks a polygon.

**Expected output:** An interactive HTML widget (in RStudio Viewer or browser) of the Africa map that you can zoom and pan with clickable pop-ups.

------------------------------------------------------------------------

## Overlay Multiple Layers {#sec-overlay}

**Purpose:** Combine disease polygons with point locations (e.g., cases, facilities) to reveal spatial relationships.

```{r overlay-layers}
set.seed(42)

# Simulate 70 health facility points across Africa for demonstration
facility_points <- st_sample(africa_health, size = 70, type = "random") |> 
  st_as_sf() |> 
  mutate(facility_id = paste0("FAC_", row_number()),
         type = "Health Facilities")

# Multi-layer map
overlay_map <- tm_shape(africa_health) +
  tm_polygons(
    "incidence", 
    fill.scale = tm_scale_continuous(values ="brewer.yl_or_rd"), 
    fill.legend = tm_legend(title = "Incidence",
                            orientation = 'portrait'),
    col = "grey60", lwd = 0.5) +
  tm_shape(facility_points) +
  tm_dots(
    fill = "type",
    fill.scale = tm_scale_categorical(values ="darkblue"),
    fill.legend = tm_legend(title = "Facilities"),
    size = 0.5
  ) +
  # tm_add_legend(
  #    type = "symbols", 
  #    labels = "Health Facilities", 
  #    fill = "darkblue", 
  #    shape = 16, # Standard round solid dot
  #    size = 0.5, # Slightly scaled up visually just for the legend box
  #    title = "Facilities"
  #  ) +
  tm_title(
    "Disease Burden & Health Facility Locations", size = 1.2
  ) +
  tm_layout(
    component.autoscale = FALSE
  )

overlay_map
```

**Key parameters:** - `tm_shape()` can be called multiple times; each subsequent layer refers to the most recent shape.
- `tm_dots()` controls point symbology.

**Expected output:** A single map showing incidence fills overlaid with blue facility dots.

------------------------------------------------------------------------

## Map Layout and Export {#sec-export}

**Purpose:** Arrange multiple maps side-by-side and save publication-quality figures.

```{r map-layout-export}
# Create two thematic maps
# Map 1
map_inc <- tm_shape(africa_health) +
  tm_fill(
    "incidence", 
    fill.scale = tm_scale(values = "brewer.yl_or_rd"),
    fill.legend = tm_legend(title = "Incidence")
  ) +
  tm_borders() +
  tm_title("A. Incidence")

map_male <- tm_shape(africa_health) +
  tm_fill(
    "male", 
    fill.scale = tm_scale(values = "brewer.blues"), 
    fill.legend = tm_legend(title = "Male Cases")
  ) +
  tm_borders() +
  tm_title(text = "B. Male Counts")

# Arrange in a grid using  tmap_arrange
combined <- tmap_arrange(map_inc, map_male, ncol = 2)
combined

# mlspatial can also be used to combine the plots
# combined <- mlspatial::plot_map_grid(list(map_inc, map_male), ncol = 2)
# combined

# Export static tmap
tmap_save(
  combined, 
  filename = "africa_disease_panel.png", 
  width = 12, height = 6, dpi = 300
)

# Export ggplot object
gg <- ggplot(africa_health) +
  geom_sf(aes(fill = incidence)) +
  scale_fill_distiller(palette = "OrRd", direction = 1)
gg

ggsave("africa_incidence_ggplot.pdf", plot = gg, width = 10, height = 8)
```

**Key parameters:** - `plot_map_grid()`: `ncol` sets the number of columns.
- `tmap_save()`: DPI and dimensions control print resolution.
- `ggsave()`: Supports `.png`, `.pdf`, `.svg`, `.tiff`.

**Expected output:** A 2-panel figure file saved to your working directory.

------------------------------------------------------------------------

## Module 2 Checklist {#sec-m2-check .unnumbered}

- [ ] Created a quick thematic map with `plot_single_map()`
- [ ] Added labels with `tm_text()`
- [ ] Built a `ggplot2` choropleth with `geom_sf()`
- [ ] Switched to interactive mode and back
- [ ] Overlaid polygon and point layers
- [ ] Exported a multi-panel figure

**Next:** Proceed to [Module 3: Spatial Autocorrelation and Cluster Analysis](03-autocorrelation.qmd).
