---
editor: 
  markdown: 
    wrap: sentence
---

# Module 2: Disease Mapping in R {#sec-module2}

This module teaches you to create publication-quality thematic maps using `mlspatial`, `tmap`, and `ggplot2`.
You will learn static maps, interactive maps, multi-layer overlays, and export techniques.

**What you will learn**

- How to build a choropleth (thematic) map in a single line with `mlspatial`.
- How `tmap` assembles a map from stacked *layers*.
- How to reproduce and customise the same maps with `ggplot2`.
- How to switch between static and interactive (Leaflet) maps.
- How to overlay point data on polygons and export multi-panel figures.

A **choropleth** map colours each region according to a data value (here, disease incidence).
It is the single most common way to visualise spatial variation in health outcomes, so this module focuses on getting the colour scale, legend, and labels right.

```{r setup-m2}
#| include: false
library(sf)
library(terra)
library(tmap)
library(ggplot2)
library(dplyr)
library(mlspatial)

data(africa_shp, package = "mlspatial")
data(panc_incidence, package = "mlspatial")
africa_health <- mlspatial::join_data(
  africa_shp |>
    mutate(
      NAME       = gsub("Côte d'Ivoire", "Cote d'Ivoire", NAME),
      COUNTRYAFF = gsub("Côte d'Ivoire", "Cote d'Ivoire", COUNTRYAFF)
    ),
  panc_incidence,
  by = "NAME"
)


# Preparing the data (recap)
# 
# As in Module 1, we first align the country names — the shapefile writes "Côte d'Ivoire" with an accent, the health table without — so the join does not silently drop the country. gsub() performs that substitution, and join_data() attaches the health columns to the polygons. The result africa_health is an sf object that is now ready to map.
```

------------------------------------------------------------------------

## Quick Thematic Maps with `mlspatial` {#sec-quick-map}

**Purpose:** Generate a publication-ready choropleth map in one line.

```{r quick-map-mlspatial}
map1 <- mlspatial::plot_single_map(
  sf_data = africa_health,
  var = "incidence",
  title = "Incidence (per 100,000)",
  palette = "yl_or_rd"
)
map1
```

::: callout-note
## How `plot_single_map()` works

This convenience function hides several `tmap` steps behind one call:

- `sf_data`: the `sf` object to draw.
- `var`: the column to colour by — **quoted**, because it is passed as a string to the plotting engine.
- `title`: the legend title.
- `palette`: the colour scheme. It accepts any ColourBrewer or viridis name written in lower-case with underscores (e.g. `"yl_or_rd"`, `"blues"`).

Under the hood, `plot_single_map()` calls `tm_fill()` with a *quantile* scale.
That means it divides the incidence values into five bins, each containing an equal number of countries, and colours each bin.
This is robust to skewed data (a few very high values would otherwise compress all other colours into one bin).
:::

**Key parameters:** - `var`: Column name to map (must be quoted).
- `palette`: Any RColourBrewer or viridis palette name.
`"yl_or_rd"` (old name `"YlOrRd"`) is standard for public-health "hot" maps.

**Expected output:** A static tmap object showing African countries coloured by pancreatic cancer incidence.

------------------------------------------------------------------------

## Add Country Labels to Maps {#sec-labels}

**Purpose:** Annotate maps so stakeholders can identify regions without a separate legend.

```{r labels-qtm}
#| message: false
#| warning: false

labelled_map <- tm_shape(africa_health) +
  # Fill layer (colours the polygons)
  tm_fill(
    "incidence", 
    fill.scale = tm_scale(values = "brewer.yl_or_rd"),  
    fill.legend = tm_legend(title = "Rate per 100k")    
  ) +
  # Border layer (outlines)
  tm_borders(col = "grey50", lwd = 0.6) +
  # Text layer (adds country names)
  tm_text(
    "NAME", 
    size = 0.6, 
    col = "black"
  ) +
  # Title layer
  tm_title("Pancreatic Cancer Incidence in Africa",
           color = 'black', size = 1.2) + 
  # Layout (everything else about map appearance)
  tm_layout(
    legend.position = c("left", "bottom")
  ) +
  # Add compass symbol
  tm_compass(
    type = "8star", 
    position = c("right", "bottom")
  ) +
  # Add scale bar 
  tm_scalebar(
    position = c("right", "bottom")
  )
labelled_map


# You can also create a quick map using the qtm() function (less flexible)
# qtm(africa_health, fill = "incidence", fill.scale = tm_scale(values = "brewer.yl_or_rd"))
```

::: callout-note
## The `tmap` layer grammar

`tmap` builds maps the way `ggplot2` builds plots: by **adding layers** with `+`.
Each `tm_*()` function adds one element:

- `tm_shape(data)` declares the data for the layers that follow it. It does not draw anything by itself.
- `tm_fill("incidence")` colours each polygon according to `incidence`.
- `tm_borders()` draws the polygon outlines.
- `tm_text("NAME")` places the value of the `NAME` column at each polygon's centroid.
- `tm_title()`, `tm_compass()`, `tm_scalebar()`, and `tm_layout()` add map furniture (title, north arrow, scale bar, and general layout).

In `tmap` version 4, styling is split into two arguments: `*.scale` controls **how values map to visuals** (colour palette, breaks), and `*.legend` controls **how the legend looks** (title, position).
So `fill.scale = tm_scale(values = "brewer.yl_or_rd")` sets the palette, while `fill.legend = tm_legend(title = "Rate per 100k")` sets the legend title.
:::

**Key parameters:** - `tm_text()`: `text = "NAME"` specifies the label column.
- `size` is in map units; adjust downward for small countries.

**Expected output:** A choropleth map with country names overlaid at centroids.

------------------------------------------------------------------------

## Create Disease Distribution Choropleth Maps with `ggplot2` {#sec-ggplot}

**Purpose:** Leverage the tidyverse grammar of graphics for highly customized static maps.

```{r ggplot-choropleth}
library(ggplot2)

ggplot(africa_health) +
  geom_sf(aes(fill = incidence), colour = "white", linewidth = 0.2) +
  scale_fill_distiller(
    palette = "YlOrRd", direction = 1,
    name = "Incidence",
    breaks = c(5, 2000, 4000)
  ) +
  theme_minimal() +
  labs(
    title = "Pancreatic Cancer Incidence in Africa",
    subtitle = "Age-standardized rate, 2021 estimates",
    caption = "Source: GBD 2021 via mlspatial",
    x = "", y = ""
  ) +
  # geom_text(
  #   aes(label = stringr::str_wrap(NAME, 8), geometry = geometry),
  #   stat = "sf_coordinates",
  #   colour = 'black', size = 3
  # ) +
  geom_sf_text(
    aes(label = stringr::str_wrap(NAME, 8)),
    size = 3.2, colour = 'black', 
    check_overlap = TRUE  # avoids overlapping labels
  ) +
  theme(
    legend.position = "bottom",
    plot.title = element_text(face = "bold", size = 14)
  ) +
  guides(
    fill = guide_colourbar(barheight = 0.5, barwidth = 8)
  )
```

::: callout-note
## The grammar of graphics, mapped

`ggplot2` builds a plot from three ideas — **data**, **aesthetics**, and **geometries** — layered with `+`:

- `ggplot(africa_health)` supplies the data.
- `geom_sf()` is the geometry for spatial data. Inside `aes(fill = incidence)` we *map* the `incidence` column to the fill colour: each country gets a colour according to its value.
- `colour = "white"` and `linewidth = 0.2` are set **outside** `aes()`, meaning they are fixed values (thin white borders) rather than mapped to data.
- `scale_fill_distiller()` controls the fill *scale*: the `YlOrRd` palette, `direction = 1` (light-to-dark), the legend `name`, and `breaks` (where tick labels appear on the legend).
- `geom_sf_text()` adds labels. `stringr::str_wrap(NAME, 8)` inserts line breaks so long names fit, and `check_overlap = TRUE` drops labels that would collide.
- `theme()` and `guides()` fine-tune appearance (legend position and title size).
:::

**Key parameters:** - `geom_sf()`: The workhorse for sf objects.
`aes(fill = incidence)` maps the variable to colour.
- `scale_fill_distiller()`: For continuous data.
Perceptually uniform, colour-blind safe, and prints well in greyscale.

**Expected output:** A static ggplot map with a continuous colour legend.

------------------------------------------------------------------------

## Create Interactive Maps with `tmap` {#sec-interactive}

**Purpose:** Allow panning, zooming, and layer toggling for exploratory data analysis.

```{r interactive-tmap}
#| eval: true
#| message: false

# Switch tmap to interactive mode
tmap_mode("view")

interactive_map <- tm_shape(africa_health) +
  tm_fill(
    "incidence", 
    fill.scale = tm_scale(values = "brewer.yl_or_rd"), 
    fill.legend = tm_legend(title = "Incidence (per 100k)"),
    popup.vars = c("incidence", "female", "male")
  ) +
  tm_borders(col = "black", lwd = 0.5)

interactive_map

# Return to static mode for publication figures
tmap_mode("plot")
```

::: callout-note
## Static vs interactive

`tmap` has two rendering modes controlled by `tmap_mode()`:

- `"plot"` (default) draws a static image, ideal for documents and publication.
- `"view"` renders the same map as a **Leaflet** JavaScript widget that you can pan, zoom, and click.

The map code itself is identical in both modes — you only change the mode.
`popup.vars` lists the columns shown in a pop-up when a user clicks a polygon, which is only relevant in interactive mode.
The chunk is marked `eval: false` so the HTML widget is not embedded in the book; run it in RStudio to interact with it.
:::

**Key parameters:** - `tmap_mode("view")`: Renders to a Leaflet JavaScript widget.
- `popup.vars`: Vector of column names shown when a user clicks a polygon.

**Expected output:** An interactive HTML widget (in RStudio Viewer or browser) of the Africa map that you can zoom and pan with clickable pop-ups.

------------------------------------------------------------------------

## Overlay Multiple Layers {#sec-overlay}

**Purpose:** Combine disease polygons with point locations (e.g., cases, facilities) to reveal spatial relationships.

```{r overlay-layers}
set.seed(4607)

# Simulate 70 health facility points across Africa for demonstration
facility_points <- st_sample(africa_health, size = 70, type = "random") |> 
  st_as_sf() |> 
  mutate(facility_id = paste0("FAC_", row_number()),
         type = "Health Facilities")

# Multi-layer map
overlay_map <- tm_shape(africa_health) +
  tm_polygons(
    "incidence", 
    fill.scale = tm_scale_continuous(values = "brewer.yl_or_rd"), 
    fill.legend = tm_legend(title = "Incidence",
                            orientation = 'portrait'),
    col = "grey60", lwd = 0.5) +
  tm_shape(facility_points) +
  tm_dots(
    fill = "type",
    fill.scale = tm_scale_categorical(values = "darkblue"),
    fill.legend = tm_legend(title = "Facilities"),
    size = 0.5
  ) +
  # tm_add_legend(
  #    type = "symbols", 
  #    labels = "Health Facilities", 
  #    fill = "darkblue", 
  #    shape = 16, # Standard round solid dot
  #    size = 0.5, # Slightly scaled up visually just for the legend box
  #    title = "Facilities"
  #  ) +
  tm_title(
    "Disease Burden & Health Facility Locations", size = 1.2
  ) +
  tm_layout(
    component.autoscale = FALSE
  )

overlay_map
```

::: callout-note
## Simulating points and stacking layers

- `set.seed(4607)` makes the random points reproducible — re-running the chunk gives the same 70 points every time.
- `st_sample(africa_health, size = 70, type = "random")` places 70 random points *inside* the polygons. `st_as_sf()` wraps them as an `sf` object, and `mutate()` adds an ID column (`paste0("FAC_", row_number())`) and a category label.
- The map then calls `tm_shape()` **twice**: first for the polygons, then for the points. Each `tm_shape()` switches the active data, so the layers that follow it draw that data. `tm_polygons()` fills and outlines the countries; `tm_dots()` draws the points as filled circles.
- `tm_scale_categorical(values = "darkblue")` is used because `type` is a categorical variable (all points are "Health Facilities"), so a single colour — not a ramp — is appropriate.
:::

**Key parameters:** - `tm_shape()` can be called multiple times; each subsequent layer refers to the most recent shape.
- `tm_dots()` controls point symbology.

**Expected output:** A single map showing incidence fills overlaid with blue facility dots.

------------------------------------------------------------------------

## Map Layout and Export {#sec-export}

**Purpose:** Arrange multiple maps side-by-side and save publication-quality figures.

```{r map-layout-export}
#| message: false
#| warning: false

# Create two thematic maps
# Map 1
map_inc <- tm_shape(africa_health) +
  tm_fill(
    "incidence", 
    fill.scale = tm_scale(values = "brewer.yl_or_rd"),
    fill.legend = tm_legend(title = "Incidence")
  ) +
  tm_borders() +
  tm_title("A. Incidence")

map_male <- tm_shape(africa_health) +
  tm_fill(
    "male", 
    fill.scale = tm_scale(values = "brewer.blues"), 
    fill.legend = tm_legend(title = "Male Cases")
  ) +
  tm_borders() +
  tm_title(text = "B. Male Counts")

# Arrange in a grid using tmap_arrange
combined <- tmap_arrange(map_inc, map_male, ncol = 2)
combined

# mlspatial can also be used to combine the plots
# combined <- mlspatial::plot_map_grid(list(map_inc, map_male), ncol = 2)
# combined

# Export static tmap
tmap_save(
  combined, 
  filename = "africa_disease_panel.png", 
  width = 12, height = 6, dpi = 300
)

# Export ggplot object
gg <- ggplot(africa_health) +
  geom_sf(aes(fill = incidence)) +
  scale_fill_distiller(palette = "OrRd", direction = 1)
gg

ggsave("africa_incidence_ggplot.pdf", plot = gg, width = 10, height = 8)
```

::: callout-note
## Combining and saving figures

- `tmap_arrange(map_inc, map_male, ncol = 2)` places two (or more) tmap objects side-by-side in a single figure. `mlspatial::plot_map_grid()` is a thin wrapper over it.
- `tmap_save()` writes a tmap figure to disk. `width`/`height` are in inches and `dpi` (dots per inch) controls resolution — 300 dpi is the usual print standard.
- `ggsave()` is the `ggplot2` equivalent and infers the format from the file extension (`.png`, `.pdf`, `.svg`, `.tiff`, etc.).
:::

**Key parameters:** - `plot_map_grid()`: `ncol` sets the number of columns.
- `tmap_save()`: DPI and dimensions control print resolution.
- `ggsave()`: Supports `.png`, `.pdf`, `.svg`, `.tiff`.

**Expected output:** A 2-panel figure file saved to your working directory and a pdf file containing a single map.

------------------------------------------------------------------------

## Practical Exercise {#sec-m2-exercise}

Apply the mapping tools to a different variable and check your answers against the solution.

1.  Use `plot_single_map()` to map the `male` variable with the `"blues"` palette and add an appropriate title.
2.  Recreate that map in `ggplot2` with `geom_sf()`, using a different palette from `scale_fill_distiller()`, and add a subtitle.
3.  Build a `tmap` choropleth of `female` that includes borders, a legend title, a compass, and a scale bar.
4.  *(Challenge)* Simulate 50 random points with `st_sample()` (use your own seed), overlay them on an incidence map with `tm_dots()`, and give them a single colour.
5.  Arrange your two `tmap` maps (incidence and male) side-by-side and save them as a PNG at 250 dpi.

::: {.callout-note collapse="true"}
## Model solution

``` r
library(sf)
library(tmap)
library(ggplot2)
library(dplyr)
library(mlspatial)

data(africa_shp, package = "mlspatial")
data(panc_incidence, package = "mlspatial")
africa_health <- mlspatial::join_data(
  africa_shp |>
    mutate(NAME = gsub("Côte d'Ivoire", "Cote d'Ivoire", NAME)),
  panc_incidence, by = "NAME"
)

# 1. Quick map of male counts
mlspatial::plot_single_map(
  sf_data = africa_health,
  var = "male",
  title = "Male Cases",
  palette = "blues"
)

# 2. ggplot version with a subtitle
ggplot(africa_health) +
  geom_sf(aes(fill = male), colour = "white", linewidth = 0.2) +
  scale_fill_distiller(palette = "Blues", direction = 1) +
  labs(title = "Male Pancreatic Cancer Cases",
       subtitle = "GBD 2021 estimates") +
  theme_minimal()

# 3. tmap choropleth of female with furniture
tm_shape(africa_health) +
  tm_fill("female",
          fill.scale = tm_scale(values = "brewer.purples"),
          fill.legend = tm_legend(title = "Female Cases")) +
  tm_borders(col = "grey40", lwd = 0.5) +
  tm_compass(position = c("right", "top")) +
  tm_scalebar(position = c("right", "bottom"))

# 4. Random points overlaid (own seed)
set.seed(5327)
pts <- st_sample(africa_health, size = 50, type = "random") |>
  st_as_sf() |>
  mutate(type = "Facilities")

tm_shape(africa_health) +
  tm_polygons("incidence",
              fill.scale = tm_scale_continuous(values = "brewer.yl_or_rd"),
              fill.legend = tm_legend(title = "Incidence"),
              col = "grey60", lwd = 0.5) +
  tm_shape(pts) +
  tm_dots(fill = "type",
          fill.scale = tm_scale_categorical(values = "darkblue"),
          fill.legend = tm_legend(title = "Facilities"),
          size = 0.4)

# 5. Arrange and save
map_inc <- tm_shape(africa_health) +
  tm_fill("incidence",
          fill.scale = tm_scale(values = "brewer.yl_or_rd"),
          fill.legend = tm_legend(title = "Incidence")) +
  tm_borders() + tm_title("A. Incidence")

map_male <- tm_shape(africa_health) +
  tm_fill("male",
          fill.scale = tm_scale(values = "brewer.blues"),
          fill.legend = tm_legend(title = "Male Cases")) +
  tm_borders() + tm_title("B. Male Counts")

combined <- tmap_arrange(map_inc, map_male, ncol = 2)
tmap_save(combined, filename = "exercise_panel.png",
          width = 12, height = 6, dpi = 150)
```
:::

------------------------------------------------------------------------

## Module 2 Checklist {#sec-m2-check .unnumbered}

- [ ] Created a quick thematic map with `plot_single_map()`
- [ ] Added labels with `tm_text()`
- [ ] Built a `ggplot2` choropleth with `geom_sf()`
- [ ] Switched to interactive mode and back
- [ ] Overlaid polygon and point layers
- [ ] Exported a multi-panel figure
- [ ] Practical exercise completed

**Next:** Proceed to [Module 3: Spatial Autocorrelation and Cluster Analysis](03-autocorrelation.qmd).
