---
editor: 
  markdown: 
    wrap: 72
---

# Module 3: Spatial Autocorrelation and Cluster Analysis {#sec-module3}

This module introduces the concept of spatial autocorrelation—whether
disease rates in one location are similar to rates in nearby locations.
You will build spatial weights matrices, compute global and local
Moran's I, detect hotspots with Getis-Ord Gi\*, and visualize the
results.

```{r setup-m3}
#| include: false
library(sf)
library(spdep)
library(dplyr)
library(mlspatial)
library(tmap)

data(africa_shp, package = "mlspatial")
data(panc_incidence, package = "mlspatial")
africa_health <- join_data(
  africa_shp |> mutate(
    across(c(NAME, COUNTRYAFF),
           ~ replace_values(.x, "Côte d'Ivoire" ~ "Cote d'Ivoire"))),
  panc_incidence, by = "NAME"
)
africa_health <- st_make_valid(africa_health)
```

------------------------------------------------------------------------

## Build Spatial Weights Matrix {#sec-weights}

**Purpose:** Define which polygons are "neighbors" so we can test
whether disease rates in one country resemble those in adjacent
countries.

> **Tobler's First Law of Geography:** *Everything is related to
> everything else, but near things are more related than distant
> things.*

```{r spatial-weights}
# Create adjacency neighbor list (queen contiguity: shares any boundary point) using spdep package
nb <- poly2nb(
  africa_health, 
  row.names = africa_health$NAME, 
  queen = TRUE,
  snap = 1000
)

# Convert neighbor list to spatial weights list
# style = "W" = row-standardized (each row sums to 1)
# zero.policy = TRUE allows regions with no neighbors (e.g., islands)
lw <- nb2listw(nb, style = "W", zero.policy = TRUE)

# Summary of neighbor connections
summary(nb)
```

**Key parameters:** - `poly2nb()`: `queen = TRUE` defines neighbours if
they share even a single point; `queen = FALSE` (rook) requires a shared
edge. - `nb2listw()`: `style = "W"` row-standardizes; `"B"` is binary
(0/1); `"C"` is globally standardized.

**Expected output:** Summary showing average number of links, number of
regions with 0 neighbors, etc.

------------------------------------------------------------------------

## Calculate Global Moran's I {#sec-global-moran}

**Purpose:** Test whether there is **overall** spatial clustering in the
disease rate across the entire study area.

```{r global-moran}
# Method 1: Direct spdep (classic approach, recommended)
global_moran <- moran.test(
  x = africa_health$incidence,
  listw = lw,
  zero.policy = TRUE,
  na.action = na.exclude
)
print(global_moran)

# Method 2: mlspatial wrapper (computes global + local in one call)
# autocorr_result <- compute_spatial_autocorr(
#   sf_data = africa_health,
#   values = africa_health$incidence,
#   signif = 0.05
# )
# print(autocorr_result$moran)
```

**Key parameters:**

- `moran.test()`: Tests the null hypothesis of spatial randomness. -
  Statistic range: −1 (perfect dispersion) to +1 (perfect clustering); 0
  = random.

**Expected output:**

\- Moran's I statistic: ≈ +1: strong positive autocorrelation (high
value near high, low near low), ≈ 0: No spatial autocorrelation (random
pattern) and ≈ -1: Strong negative autocorrelation (high near low, like
a checkerboard pattern)

\- p value \< 0.05 indicates significant clustering and vise versa.

------------------------------------------------------------------------

## Calculate Local Moran's I (LISA) {#sec-local-moran}

**Purpose:** Identify **which specific locations** contribute to the
global clustering (local clusters). LISA stands for local indicators of
Spatial Association.

```{r local-moran}
# Method 1: spdep localmoran
local_mi <- localmoran(
  x = africa_health$incidence,
  listw = lw,
  zero.policy = TRUE,
  na.action = na.exclude
)
head(local_mi)

# Attach results to sf object
africa_health$local_i <- local_mi[, "Ii"]
africa_health$local_i_p <- local_mi[, "Pr(z != E(Ii))"]
head(africa_health)

# Method 2: mlspatial (recommended for integrated workflow)
# The wrapper already added standardized values, spatial lag, local Moran's I,
# z-scores, p-values, and cluster classification to autocorr_result$data
# lisa_sf <- autocorr_result$data
# head(lisa_sf)
```

**Key parameters:** - \``local_moran()` output is a matrix of the local
statistics. `Ii` tells the strength and direction of local pattern,
`E.Ii` tells the baseline under randomness, `Var.Ii` shows the estimate
uncertainties (variance). `Z.Ii` is the z-score (standardized deviation)
and `Pr(z != E(Ii))` is the p-value.

**Expected output:** - `localmoran()`: Returns `Ii` (local statistic),
`E.Ii`(expected statistic), variance, z-score, and p-value for each
polygon. - the `compute_spatial_autocorr()` from `mlspatial` generates
the local Moran statistics in addition to the global Moran's.

------------------------------------------------------------------------

## Hotspot Analysis (`Getis-Ord Gi*`) {#sec-gi}

**Purpose:** Distinguish statistically significant **hotspots** (high
values surrounded by high values) from **coldspots** (low-low) using the
$Getis-Ord\ Gi^*$ statistic.

```{r hotspot-gi}
# localG returns z-scores; positive = hotspot, negative = coldspot
# Note: localG does not allow zero-policy neighbours by default, 
# so we first add country to its own neigbour list (Useful for Gi*)
nb_self <- include.self(nb)
summary(nb_self)

gi_star <- localG(
  x = africa_health$incidence,
  listw = nb2listw(nb_self, style='W', zero.policy = TRUE),
  zero.policy = TRUE
)

# Add Gi* to dataset
africa_health <- africa_health |> 
  mutate(gi_star = as.numeric(gi_star))

# Classify hotspots at alpha = 0.05 (|z| > 1.96)
# Method 1: Using standard normal distribution sutoff
africa_health <- africa_health |> 
  mutate(hotspot = case_when(
  gi_star > 1.96 ~ "Hotspot (High-High)",
  gi_star < -1.96 ~ "Coldspot (Low-Low)",
  TRUE ~ "Not significant"
 ))

# Method 2: Using the exact p-values from localG (recommended)
gi_internals <- attr(gi_star, "internals")

africa_health$gi_z <- gi_internals[, "Z(G*i)"]
africa_health$gi_p <- gi_internals[, "Pr(z != E(G*i))"]

africa_health <- africa_health |> 
  mutate(
    hotspot_exact = case_when(
    gi_z > 0 & gi_p <= 0.05 ~ "Hotspot",
    gi_z < 0 & gi_p <= 0.05 ~ "Coldspot",
    TRUE ~ "Not significant"),
    hotspot_exact = factor(hotspot_exact,
                           levels = c("Coldspot", "Hotspot", "Not significant"))
  )


table(africa_health$hotspot_exact)
```

**Key parameters:** -`include.self()`: Adds exactly one neighbour to
each country (itself). This helps the Gi\* calculation since it works by
calculating a local sum for each country (its value plus neighbours
value). It essentially shifts the whole distribution by +1. `localG()`:
Computes $Gi^*$ which includes the focal unit in its own neighborhood.
The values calculated are Z scores. - z \> 1.96 indicates a
statistically significant hotspot; z \< −1.96 indicates a coldspot and z
near 0 is not significant.

**Expected output:** A frequency table showing how many countries are
classified as hotspots, coldspots, or non-significant.

------------------------------------------------------------------------

## Visualize Cluster Results {#sec-visualize-clusters}

**Purpose:** Map LISA clusters and Gi\* hotspots to communicate findings
to decision-makers. The local MI values contains an attribute called
`quadr`. We can use this to visualise the local clusters.

```{r visualize-clusters}
# Map LISA clusters from mlspatial
# extract the quadrant and exact p-values from the local_mi
africa_health$lisa_quadrant <- attributes(local_mi)$quadr$mean

#filter the quadrants by the 0.05 significance threshold
africa_health <- africa_health |> 
  mutate(
    lisa_cluster = case_when(
    local_i_p <= 0.05 ~ as.character(lisa_quadrant),
    TRUE ~ "Not significant"),
    lisa_cluster = factor(lisa_cluster,
                          levels = c("Low-Low", "High-Low", "Low-High", "High-High", "Not significant"))
  )

# Create map
lisa_colours <- c(
  "High-High" = "red", 
  "Low-Low" = "blue", 
  "High-Low" = "lightpink", 
  "Low-High" = "lightblue", 
  "Not significant" = "grey90"
)
lisa_map <- tm_shape(africa_health) +
  tm_polygons(
    fill = "lisa_cluster",
    fill.scale = tm_scale_categorical(values = lisa_colours),
    fill.legend = tm_legend(title = "LISA Clusters (Pancreatic Incidence)"),
    col = "black",
    col_alpha = 0.8,
    lwd = 0.5
  ) +
  tm_title("Local Moran's I (LISA) Cluster Map")

# Map Gi* hotspots
Gi_colours <- c(
  "Hotspot" = "red",
  "Coldspot" = "blue",
  "Not significant" = 'grey90'
)
hotspot_map <- tm_shape(africa_health) +
  tm_polygons(
    fill = "hotspot_exact", 
    fill.scale = tm_scale(values = Gi_colours),
    fill.legend = tm_legend(title = "Gi* Classification")
  ) +
  tm_title("Getis-Ord Gi* Hotspots")

hotspots <- tmap_arrange(lisa_map, hotspot_map, ncol =2)
hotspots

tmap_save(
  hotspots,
  filename = "africa_hotspot_panel.png", 
  width = 12, height = 6, dpi = 300
)
```

**Expected output:** A two-panel map. The left panel shows LISA
clusters; the right panel shows `Gi*` hotspots and coldspots.

------------------------------------------------------------------------

## Module 3 Checklist {#sec-m3-check .unnumbered}

- [ ] Built a spatial weights matrix with `poly2nb()` and `nb2listw()`
- [ ] Computed Global Moran's I and interpreted the p-value
- [ ] Computed Local Moran's I (LISA) and identified cluster types
- [ ] Calculated Getis-Ord Gi\* and classified hotspots/coldspots
- [ ] Visualized both LISA and Gi\* results side-by-side

**Next:** Proceed to [Module 4: Spatial Regression and Environmental
Factor Analysis](04-regression.qmd).
