---
editor: 
  markdown: 
    wrap: 72
---

# Module 3: Spatial Autocorrelation and Cluster Analysis {#sec-module3}

This module introduces the concept of spatial autocorrelation—whether
disease rates in one location are similar to rates in nearby locations.
You will build spatial weights matrices, compute global and local
Moran's I, detect hotspots with `Getis-Ord Gi*`, and visualize the
results.

**What you will learn**

- What spatial autocorrelation is and why it matters in epidemiology.
- How to define "neighbours" and build a spatial weights matrix.
- How to test for global clustering with Moran's I.
- How to locate *where* clustering happens with Local Moran's I (LISA).
- How to find hotspots/coldspots with `Getis-Ord Gi*.`

> **Tobler's First Law of Geography:** *Everything is related to
> everything else, but near things are more related than distant
> things.* Spatial autocorrelation is the formal way to measure whether
> this is true for your data.

```{r setup-m3}
#| include: false
library(sf)
library(spdep)
library(dplyr)
library(mlspatial)
library(tmap)

data(africa_shp, package = "mlspatial")
data(panc_incidence, package = "mlspatial")
africa_health <- join_data(
  africa_shp |> mutate(
    NAME       = gsub("Côte d'Ivoire", "Cote d'Ivoire", NAME),
    COUNTRYAFF = gsub("Côte d'Ivoire", "Cote d'Ivoire", COUNTRYAFF)
  ),
  panc_incidence, by = "NAME"
)
africa_health <- st_make_valid(africa_health)
```

::: callout-note
## Why spatial autocorrelation matters

If a disease were randomly distributed, every region's rate would be
independent of its neighbours. In reality, infectious diseases,
environmental risk factors, and health-service access all spill across
borders, so nearby regions tend to resemble each other. Testing for
spatial autocorrelation tells you whether that resemblance is stronger
than chance — and, if so, *where* the clusters are. This matters for two
reasons: it reveals real disease hotspots, and it flags a violation of
the independence assumption behind ordinary (non-spatial) regression
models, which Module 4 addresses.
:::

------------------------------------------------------------------------

## Build Spatial Weights Matrix {#sec-weights}

**Purpose:** Define which polygons are "neighbours" so we can test
whether disease rates in one country resemble those in adjacent
countries.

Every spatial statistic needs a definition of **"who is near whom"**.
That definition is stored in a *spatial weights matrix*: for each
region, a list of its neighbours and the weight given to each.

```{r spatial-weights}
# Create adjacency neighbour list (queen contiguity: shares any boundary point)
nb <- poly2nb(
  africa_health, 
  row.names = africa_health$NAME, 
  queen = TRUE,
  snap = 1000
)

# Convert neighbour list to spatial weights list
# style = "W" = row-standardized (each row sums to 1)
# zero.policy = TRUE allows regions with no neighbors (e.g., islands)
lw <- nb2listw(nb, style = "W", zero.policy = TRUE)

# Summary of neighbour connections
summary(nb)
```

::: callout-note
## Reading the weights-matrix code

- `poly2nb()` builds the *neighbour list*. For polygon data the usual
  rule is **contiguity**: two regions are neighbours if they touch.
  `queen = TRUE` (queen contiguity) treats them as neighbours if they
  share any point or edge, like chess-queen moves; `queen = FALSE` (rook
  contiguity) requires a shared *edge*.
- `snap = 1000` is a distance tolerance in the units of your data.
  Because `africa_health` is projected in metres (EPSG:3857), this
  tolerates gaps up to 1,000 m between supposedly adjacent borders — a
  common fix for boundaries that don't quite line up.
- `row.names` attaches the country name to each entry, making the object
  easier to read.
- `nb2listw()` converts the neighbour list into *weights*. `style = "W"`
  row-standardizes, so each neighbour's weight is
  `1 / (number of neighbours)` and every row sums to 1. Alternatives are
  `"B"` (binary 0/1) and `"C"` (globally standardized).
- `zero.policy = TRUE` is essential here: island countries have no
  contiguous neighbours, and without this flag the functions would error
  on them.
:::

**Key parameters:** - `poly2nb()`: `queen = TRUE` defines neighbours if
they share even a single point; `queen = FALSE` (rook) requires a shared
edge. - `nb2listw()`: `style = "W"` row-standardizes; `"B"` is binary
(0/1); `"C"` is globally standardized.

**Expected output:** The `summary(nb)` reports 54 regions connected by
214 links (an average of about 4 neighbours each), with **6 regions
having no neighbours** — the island nations Cabo Verde, Comoros, São
Tomé and Príncipe, Madagascar, Mauritius, and Seychelles. There are 7
disconnected sub-graphs (the mainland plus six islands).

------------------------------------------------------------------------

## Calculate Global Moran's I {#sec-global-moran}

**Purpose:** Test whether there is **overall** spatial clustering in the
disease rate across the entire study area.

Global Moran's I summarizes the whole map with a single number: are
similar values generally near each other?

```{r global-moran}
# Method 1: Direct spdep (classic approach, recommended)
global_moran <- moran.test(
  x = africa_health$incidence,
  listw = lw,
  zero.policy = TRUE,
  na.action = na.exclude
)
print(global_moran)

# Method 2: mlspatial wrapper (computes global + local in one call)
# autocorr_result <- compute_spatial_autocorr(
#   sf_data = africa_health,
#   values = africa_health$incidence,
#   signif = 0.05
# )
# print(autocorr_result$moran)
```

::: callout-note
## Interpreting Moran's I

Moran's I ranges from **−1 to +1**:

- **+1** — perfect clustering: high values cluster with high, low with
  low.
- **0** — no spatial autocorrelation: values are arranged randomly.
- **−1** — perfect dispersion: high values sit next to low ones (a
  "checkerboard" pattern).

`moran.test()` compares the observed I (Index) against a null hypothesis
of spatial randomness and returns a p-value. A p-value below 0.05 means
the clustering is unlikely to be due to chance.

**In this dataset the observed I is about −0.07 with a p-value of about
0.75.** That is close to zero and *not* significant, so there is no
strong continent-wide clustering of incidence. This is an important
result in itself: a global statistic can be non-significant **even when
there are local clusters** (a few high-value regions cancelling out a
few low-value ones). That is exactly why we turn next to *local*
statistics.
:::

**Key parameters:**

- `moran.test()`: Tests the null hypothesis of spatial randomness. -
  Statistic range: −1 (perfect dispersion) to +1 (perfect clustering); 0
  = random.

**Expected output:**

- Moran's I statistic ≈ +1: strong positive autocorrelation (high value
  near high, low near low); ≈ 0: no spatial autocorrelation (random
  pattern); ≈ −1: strong negative autocorrelation (high near low, like a
  checkerboard pattern).

- A p-value \< 0.05 indicates significant clustering, and vice versa.

------------------------------------------------------------------------

## Calculate Local Moran's I (LISA) {#sec-local-moran}

**Purpose:** Identify **which specific locations** contribute to the
global clustering (local clusters). LISA stands for Local Indicators of
Spatial Association.

While the global statistic gives one number for the whole map, LISA
computes a statistic **per region**, so you can map where the clusters
are.

```{r local-moran}
# Method 1: spdep localmoran
local_mi <- localmoran(
  x = africa_health$incidence,
  listw = lw,
  zero.policy = TRUE,
  na.action = na.exclude
)
head(local_mi)

# Attach results to sf object
africa_health$local_i <- local_mi[, "Ii"]
africa_health$local_i_p <- local_mi[, "Pr(z != E(Ii))"]
head(africa_health)

# Method 2: mlspatial (recommended for integrated workflow)
# The wrapper already added standardized values, spatial lag, local Moran's I,
# z-scores, p-values, and cluster classification to autocorr_result$data
# lisa_sf <- autocorr_result$data
# head(lisa_sf)
```

::: callout-note
## Reading the LISA output

`localmoran()` returns a matrix with one row per region and these
columns:

- `Ii` — the local Moran statistic: its sign and size show the strength
  and direction of the local pattern.
- `E.Ii` — the expected value under spatial randomness (the baseline).
- `Var.Ii` — the variance (uncertainty) of that estimate.
- `Z.Ii` — the z-score (how many standard errors `Ii` is from `E.Ii`).
- `Pr(z != E(Ii))` — the p-value for that region.

A region with a small p-value (e.g. \< 0.05) is a *significant* local
cluster. Note that the six island countries with no neighbours receive
`NA` here — their local statistic is undefined, which is why
`zero.policy = TRUE` was needed.

The quadrant classification (High-High, Low-Low, etc.) is stored in an
attribute rather than the matrix itself; we extract it in the
visualization section below.
:::

**Key parameters:** - `local_moran()` output is a matrix of the local
statistics. `Ii` tells the strength and direction of local pattern,
`E.Ii` tells the baseline under randomness, `Var.Ii` shows the estimate
uncertainties (variance). `Z.Ii` is the z-score (standardized deviation)
and `Pr(z != E(Ii))` is the p-value.

**Expected output:** - `localmoran()`: Returns `Ii` (local statistic),
`E.Ii` (expected statistic), variance, z-score, and p-value for each
polygon. - the `compute_spatial_autocorr()` from `mlspatial` generates
the local Moran statistics in addition to the global Moran's.

------------------------------------------------------------------------

## Hotspot Analysis (`Getis-Ord Gi*`) {#sec-gi}

**Purpose:** Distinguish statistically significant **hotspots** (high
values surrounded by high values) from **coldspots** (low-low) using the
$Getis-Ord\ Gi^*$ statistic.

Where LISA asks "is this region *correlated* with its neighbours?", the
Gi\* statistic asks a subtly different question: "is this region part of
a cluster of **high** (or **low**) values?" It is the standard tool for
labelling hotspots and coldspots on a map.

```{r hotspot-gi}
# localG returns z-scores; positive = hotspot, negative = coldspot
# Note: Because we want to know if a target country combined with
# its neighbours is a hotspot or not, 
# we first add country to its own neighbour list (useful for Gi*)
nb_self <- include.self(nb)
summary(nb_self)

gi_star <- localG(
  x = africa_health$incidence,
  listw = nb2listw(nb_self, style = 'W', zero.policy = TRUE),
  zero.policy = TRUE
)

# Add Gi* to dataset
africa_health <- africa_health |> 
  mutate(gi_star = as.numeric(gi_star))

# Classify hotspots at alpha = 0.05 (|z| > 1.96)
# Method 1: Using the standard normal distribution cutoff
africa_health <- africa_health |> 
  mutate(hotspot = case_when(
    gi_star > 1.96 ~ "Hotspot (High-High)",
    gi_star < -1.96 ~ "Coldspot (Low-Low)",
    TRUE ~ "Not significant"
  ))

# Method 2: Using the exact p-values from localG (recommended)
gi_internals <- attr(gi_star, "internals")

africa_health$gi_z <- gi_internals[, "Z(G*i)"]
africa_health$gi_p <- gi_internals[, "Pr(z != E(G*i))"]

africa_health <- africa_health |> 
  mutate(
    hotspot_exact = case_when(
      gi_z > 0 & gi_p <= 0.05 ~ "Hotspot",
      gi_z < 0 & gi_p <= 0.05 ~ "Coldspot",
      TRUE ~ "Not significant"),
    hotspot_exact = factor(hotspot_exact,
                           levels = c("Coldspot", "Hotspot", "Not significant"))
  )


table(africa_health$hotspot_exact)
```

::: callout-note
## How Gi\* works

- `include.self(nb)` adds each region to its *own* neighbour list. The
  Gi\* statistic differs from LISA in that it explicitly includes the
  focal unit in the local sum — a region is a hotspot only if it is high
  **and** surrounded by highs.
- `localG()` then computes the statistic for every region. The values it
  returns are **z-scores**, standardised against the null hypothesis of
  no clustering.
- A z-score above **+1.96** marks a significant hotspot (top 2.5% of a
  normal distribution); below **−1.96** a significant coldspot; values
  in between are "not significant". Method 1 applies this 1.96 cut-off
  directly, while Method 2 uses each region's exact p-value (equivalent
  but slightly cleaner).
- `attr(gi_star, "internals")` is where `localG()` hides the detailed
  output — a matrix with columns `G*i`, `E(G*i)`, `V(G*i)`, `Z(G*i)`,
  and `Pr(z != E(G*i))`.
- The final `factor(..., levels = ...)` fixes the order of the
  categories so the legend always reads "Coldspot, Hotspot, Not
  significant" rather than alphabetical order.
:::

**Key parameters:** - `include.self()`: Adds exactly one neighbour to
each country (itself). This helps the Gi\* calculation since it works by
calculating a local sum for each country (its value plus neighbours'
values). - `localG()`: Computes $Gi^*$ which includes the focal unit in
its own neighbourhood. The values calculated are Z scores. - z \> 1.96
indicates a statistically significant hotspot; z \< −1.96 indicates a
coldspot and z near 0 is not significant.

**Expected output:** A frequency table showing how many countries are
classified as hotspots, coldspots, or non-significant. For this dataset
the result is typically **3 hotspots, 0 coldspots, and 51
non-significant** countries.

------------------------------------------------------------------------

## Visualize Cluster Results {#sec-visualize-clusters}

**Purpose:** Map LISA clusters and Gi\* hotspots to communicate findings
to decision-makers. The local MI values contain an attribute called
`quadr` (quadrant); we can use this to visualize the local clusters.

```{r visualize-clusters}
# Map LISA clusters by extracting
# the quadrant and exact p-values from the local_mi
africa_health$lisa_quadrant <- attributes(local_mi)$quadr$mean

# filter the quadrants by the 0.05 significance threshold
africa_health <- africa_health |> 
  mutate(
    lisa_cluster = case_when(
      local_i_p <= 0.05 ~ as.character(lisa_quadrant),
      TRUE ~ "Not significant"),
    lisa_cluster = factor(lisa_cluster,
                          levels = c("Low-Low", "High-Low", "Low-High", "High-High", "Not significant"))
  )

# Create map
lisa_colours <- c(
  "High-High" = "red", 
  "Low-Low" = "blue", 
  "High-Low" = "lightpink", 
  "Low-High" = "lightblue", 
  "Not significant" = "grey90"
)
lisa_map <- tm_shape(africa_health) +
  tm_polygons(
    fill = "lisa_cluster",
    fill.scale = tm_scale_categorical(values = lisa_colours),
    fill.legend = tm_legend(title = "LISA Clusters (Pancreatic Incidence)"),
    col = "black",
    col_alpha = 0.8,
    lwd = 0.5
  ) +
  tm_title("Local Moran's I (LISA) Cluster Map")

# Map Gi* hotspots
Gi_colours <- c(
  "Hotspot" = "red",
  "Coldspot" = "blue",
  "Not significant" = 'grey90'
)
hotspot_map <- tm_shape(africa_health) +
  tm_polygons(
    fill = "hotspot_exact", 
    fill.scale = tm_scale(values = Gi_colours),
    fill.legend = tm_legend(title = "Gi* Classification")
  ) +
  tm_title("Getis-Ord Gi* Hotspots")

hotspots <- tmap_arrange(lisa_map, hotspot_map, ncol = 2)
hotspots

tmap_save(
  hotspots,
  filename = "africa_hotspot_panel.png", 
  width = 12, height = 6, dpi = 300
)
```

::: callout-note
## Reading the visualization code

- `attributes(local_mi)$quadr$mean` extracts the quadrant classification
  from the `localmoran()` output. The four quadrants are named by the
  region's value and its neighbours' values: **High-High** (a hot
  cluster), **Low-Low** (a cold cluster), **High-Low** (a high region
  surrounded by lows), and **Low-High** (the inverse).
- `case_when(local_i_p <= 0.05 ~ ..., TRUE ~ "Not significant")` keeps
  only the quadrants that are statistically significant, and relabels
  the rest. (Regions whose p-value is `NA` — the islands — fall through
  to "Not significant".)
- The named colour vector `lisa_colours` maps each category to a fixed
  colour, and `tm_scale_categorical()` applies it. Giving the `factor()`
  an explicit `levels` order guarantees the legend order matches the
  colours.
- `tmap_arrange(..., ncol = 2)` places the two maps side by side, and
  `tmap_save()` writes the panel to a PNG at 300 dpi.
:::

**Expected output:** A two-panel map. The left panel shows LISA
clusters; the right panel shows `Gi*` hotspots and coldspots.

------------------------------------------------------------------------

## Practical Exercise {#sec-m3-exercise}

Re-run the whole workflow for a different variable — the male case
counts (`male`) — and check your answers against the solution.

1.  Build a **rook** contiguity neighbour list (`queen = FALSE`, keeping
    `snap = 1000`). How does the total number of links compare with the
    queen version above?
2.  Compute global Moran's I for `male` and state whether clustering is
    significant at the 0.05 level.
3.  Compute local Moran's I for `male`. How many regions have a
    significant (p ≤ 0.05) local statistic?
4.  Run Getis-Ord Gi\* on `male` and report how many hotspots and
    coldspots you find.
5.  *(Challenge)* Map the Gi\* result for `male` using `tm_polygons()`
    with your own colour scheme.

::: {.callout-note collapse="true"}
## Model solution

``` r
library(sf)
library(spdep)
library(dplyr)
library(mlspatial)
library(tmap)

data(africa_shp, package = "mlspatial")
data(panc_incidence, package = "mlspatial")
africa_health <- join_data(
  africa_shp |> mutate(NAME = gsub("Côte d'Ivoire", "Cote d'Ivoire", NAME)),
  panc_incidence, by = "NAME"
)
africa_health <- st_make_valid(africa_health)

# 1. Rook contiguity
nb_rook <- poly2nb(africa_health, queen = FALSE, snap = 1000)
summary(nb_rook)   # fewer links than the queen version (shared edges only)

# 2. Global Moran's I for male
lw_rook <- nb2listw(nb_rook, style = "W", zero.policy = TRUE)
moran.test(africa_health$male, listw = lw_rook,
           zero.policy = TRUE, na.action = na.exclude)

# 3. Local Moran's I for male
lm_male <- localmoran(africa_health$male, listw = lw_rook,
                      zero.policy = TRUE, na.action = na.exclude)
sum(lm_male[, "Pr(z != E(Ii))"] <= 0.05, na.rm = TRUE)

# 4. Getis-Ord Gi* for male
nb_self <- include.self(nb_rook)
gi_male <- localG(africa_health$male,
                  listw = nb2listw(nb_self, style = "W", zero.policy = TRUE),
                  zero.policy = TRUE)
gi_int <- attr(gi_male, "internals")
africa_health$gi_z <- gi_int[, "Z(G*i)"]
africa_health$gi_p <- gi_int[, "Pr(z != E(G*i))"]
africa_health <- africa_health |>
  mutate(hotspot = case_when(
    gi_z > 0 & gi_p <= 0.05 ~ "Hotspot",
    gi_z < 0 & gi_p <= 0.05 ~ "Coldspot",
    TRUE ~ "Not significant"))
table(africa_health$hotspot)

# 5. Map the Gi* result
cols <- c("Hotspot" = "red", "Coldspot" = "blue", "Not significant" = "grey90")
tm_shape(africa_health) +
  tm_polygons(fill = "hotspot",
              fill.scale = tm_scale_categorical(values = cols),
              fill.legend = tm_legend(title = "Gi* (male)")) +
  tm_title("Getis-Ord Gi* Hotspots — Male Cases")
```
:::

------------------------------------------------------------------------

## Module 3 Checklist {#sec-m3-check .unnumbered}

- [ ] Built a spatial weights matrix with `poly2nb()` and `nb2listw()`
- [ ] Computed Global Moran's I and interpreted the p-value
- [ ] Computed Local Moran's I (LISA) and identified cluster types
- [ ] Calculated Getis-Ord Gi\* and classified hotspots/coldspots
- [ ] Visualized both LISA and Gi\* results side-by-side
- [ ] Practical exercise completed

**Next:** Proceed to [Module 4: Spatial Regression and Environmental
Factor Analysis](04-regression.qmd).
