---
editor: 
  markdown: 
    wrap: 72
---

# Module 4: Spatial Regression and Epidemiological Risk Factor Analysis {#sec-module4}

This module connects health outcomes to lifestyle/genetic drivers. You
will simulate covariates, build a global OLS model, test its residuals
for spatial autocorrelation, and then fit a Geographically Weighted
Regression (GWR) to capture spatially varying relationships.

**What you will learn**

- How to assemble risk-factor covariates for a regression.
- How to fit and interpret an ordinary least squares (OLS) model.
- How to spot **data leakage** — a perfect fit that means the outcome
  was derived from its own predictors.
- Why we model *rates* rather than raw counts.
- How to test OLS residuals for spatial autocorrelation.
- What Geographically Weighted Regression (GWR) is, and how to judge
  whether it is actually an improvement.

```{r setup-m4}
#| include: false
library(sf)
library(terra)
library(spdep)
library(GWmodel)
library(dplyr)
library(mlspatial)
library(tmap)

data(africa_shp, package = "mlspatial")
data(panc_incidence, package = "mlspatial")
africa_health <- join_data(
  africa_shp |> mutate(
    NAME       = gsub("Côte d'Ivoire", "Cote d'Ivoire", NAME),
    COUNTRYAFF = gsub("Côte d'Ivoire", "Cote d'Ivoire", COUNTRYAFF)
  ),
  panc_incidence, by = "NAME"
)
africa_health <- st_make_valid(africa_health)

nb <- poly2nb(africa_health, row.names = africa_health$NAME,
              queen = TRUE, snap = 1000)
lw <- nb2listw(nb, style = "W", zero.policy = TRUE)
```

::: callout-note
## Where this module sits

Modules 2 and 3 mapped the disease and measured its spatial clustering.
Here we ask a causal question: *which factors drive the disease rate?*
Ordinary regression assumes each observation is independent, but Module
3 showed spatial data can violate that. The workflow below is the
standard spatial-epidemiology sequence: fit a global model, check its
residuals for spatial structure, and only then reach for a local (GWR)
model if the data justify it.
:::

------------------------------------------------------------------------

## Simulate Risk-Factor Covariates {#sec-simulate}

**Purpose:** Link health outcomes to known epidemiological risk factors
for pancreatic cancer (smoking, obesity, diabetes, alcohol use, family
history, urbanisation) for use in the spatial regression exercise.

For this tutorial, we **simulate** six covariates directly at the
country level so the code is fully runnable without downloading external
datasets. Because `africa_health` is a polygon layer where each row
already represents one country (like `incidence`, `TOTPOP`, etc.), we do
not need raster data — we can generate one random value per row and
append it directly with `mutate()`.

**In real research**, replace these random draws with actual
country-level estimates from sources such as the WHO Global Health
Observatory (smoking, obesity, diabetes prevalence), IARC GLOBOCAN
risk-factor tables, or national health surveys. These are typically
small CSV/Excel downloads, joined to `africa_health` by country ISO code
(`ISO_3DIGIT`) rather than simulated.

```{r simulate-covariates}
set.seed(898)  # reproducibility

n <- nrow(africa_health)

africa_health <- africa_health |> 
  mutate(
    # Smoking prevalence (% of adult population) — WHO African region estimates
    # typically range ~10-25%; strongest modifiable risk factor for pancreatic cancer
    smoking_prev = round(runif(n, min = 8, max = 28), 1),
    
    # Obesity prevalence (% adults, BMI >= 30) — WHO Africa region estimates
    # range roughly 5-30% depending on urbanization/country
    obesity_prev = round(runif(n, min = 5, max = 30), 1),
    
    # Type 2 diabetes prevalence (%) — longstanding diabetes is a well-established
    # pancreatic cancer risk factor; IDF Africa estimates ~3-10% typically
    diabetes_prev = round(runif(n, min = 2, max = 11), 1),
    
    # Alcohol consumption (litres pure alcohol per capita per year) —
    # heavy long-term use linked to chronic pancreatitis -> elevated risk
    alcohol_per_capita = round(runif(n, min = 0.5, max = 12), 1),
    
    # Family history / genetic predisposition proxy (% of cases with
    # first-degree relative history) — literature suggests ~5-10% of
    # pancreatic cancers have a hereditary component
    family_history_prev = round(runif(n, min = 3, max = 10), 1),
    
    # Urbanization (%) — proxy for dietary/lifestyle transition,
    # often used as a covariate for NCD risk in African epidemiology
    urbanization_pct = round(runif(n, min = 15, max = 85), 1)
  )

head(africa_health[c("NAME", "smoking_prev", "obesity_prev", 
                     "diabetes_prev", "alcohol_per_capita", 
                     "family_history_prev", "urbanization_pct")])
```

::: callout-note
## Why we simulate (and why that matters for interpretation)

- `set.seed(898)` fixes the random-number generator so the "random"
  covariates are the same every time the document is rendered —
  essential for a reproducible tutorial.
- `runif(n, min, max)` draws `n` values from a uniform distribution
  between `min` and `max`, one per country. `round(..., 1)` keeps them
  to one decimal place.
- These covariates are simulated **independently of `incidence` and of
  each other** — they are *not* designed to correlate with pancreatic
  cancer rates. This is intentional: the goal is to demonstrate the
  regression *pipeline* (data structure, model syntax, diagnostics), not
  to produce a real epidemiological finding. As you will see, this
  choice turns out to be a useful teaching device.
:::

**Key parameters:** - `runif(n, min, max)`: draws `n` random values from
a uniform distribution bounded to a realistic range for each covariate,
loosely based on published prevalence ranges across African countries.

**Expected output:** Six new columns appended to `africa_health`:

| Column | Description | Simulated range |
|------------------------|------------------------|------------------------|
| `smoking_prev` | Adult smoking prevalence (%) | 8–28% |
| `obesity_prev` | Adult obesity prevalence, BMI ≥ 30 (%) | 5–30% |
| `diabetes_prev` | Type 2 diabetes prevalence (%) | 2–11% |
| `alcohol_per_capita` | Alcohol consumption (litres pure alcohol/year) | 0.5–12 |
| `family_history_prev` | Cases with first-degree relative history (%) | 3–10% |
| `urbanization_pct` | Urban population (%) | 15–85% |

------------------------------------------------------------------------

## Build Ordinary Least Squares (OLS) Regression {#sec-ols}

**Purpose:** Establish a global (non-spatial) baseline model relating
pancreatic cancer incidence to lifestyle, genetic and demographic risk
factors.

```{r ols-model}
ols_model <- lm(
  incidence ~ smoking_prev + obesity_prev + diabetes_prev + 
    alcohol_per_capita + family_history_prev + urbanization_pct + 
    female + male + ageb + agec,
  data = africa_health
)
summary(ols_model)
```

::: callout-note
## Reading an `lm()` summary

The formula `response ~ predictor1 + predictor2 + ...` tells `lm()` what
to fit. The `summary()` then reports, for each predictor:

- **Estimate** — the expected change in the outcome for a one-unit
  increase in that predictor, holding the others fixed.
- **Std. Error / t value / Pr(\>\|t\|)** — how precisely the coefficient
  is estimated and whether it differs from zero.
- **R-squared** — the proportion of variance in the outcome explained by
  the model (0 to 1).
- **F-statistic** — an overall test that at least one predictor matters.

A positive coefficient means higher predictor → higher expected
incidence; negative means the reverse.
:::

**Key parameters:** - Formula syntax:
`response ~ predictor1 + predictor2 + ...` - Interpretation: Positive
coefficients indicate higher expected incidence with higher predictor
values.

**Expected output:** A standard `lm` summary showing coefficients,
standard errors, t-values, and p-values. R² indicates the proportion of
variance explained globally.

::: callout-warning
## A red flag: a *perfect* fit

This first model prints a warning message and returns an $R^2$ of
exactly **1** with an astronomical F-statistic. That is not a
spatial-regression breakthrough — it is a classic case of **data
leakage**. A model that fits perfectly is almost always telling you that
the outcome is mathematically derived from one or more of its own
predictors. The next section diagnoses it.
:::

------------------------------------------------------------------------

## Diagnosing Data Leakage {#sec-diagnose}

The first **OLS** model returned `R-squared: 1`, an F-statistic of
`2.36e+31` and an explicit warning — *"essentially perfect fit: summary
may be unreliable"*. A model that fits perfectly is almost always
telling you that the outcome is mathematically derived from one or more
of its own predictors. This is called **data leakage**.

**Step 1: Inspect the coefficients.** Looking at the model summary, two
variables stood out immediately:

```         
female   9.302e-02   t = 1.005e+13   p < 2e-16
male     9.302e-02   t = 1.003e+13   p < 2e-16
```

Both coefficients are identical to four decimal figures, with t-values
around $10^{13}$. T-values that enormous are a signature of the model
detecting an algebraic identity, not a strong effect.

**Step 2: Test the suspected identity directly.**

```{r test-suspect}
cor(africa_health$incidence, africa_health$female + africa_health$male)
head(africa_health$incidence / (africa_health$female + africa_health$male))
```

A correlation of 1 between `incidence` and `female + male` confirms that
`incidence` is a fixed linear function of `female + male`. Dividing one
by the other reveals the exact relationship: every single country
returns the exact constant `0.09302326`, meaning `incidence` was
computed directly from `male` and `female` (or they share a common
upstream calculation).

Including `male` and `female` as predictors makes it **mathematically
circular** — the model is not learning but rather reversing an
arithmetic formula.

**Step 3: Screen every numeric variable for leakage.**

```{r check-cor}
# Check all numeric columns systematically against incidence
numeric_cols <- africa_health |> 
  st_drop_geometry() |>                    
  select(where(is.numeric)) |> 
  select(-incidence) |>                    # exclude the outcome itself
  names()

cor_check <- sapply(numeric_cols, function(col) {
  cor(africa_health$incidence, africa_health[[col]], use = "complete.obs")
})

sort(abs(cor_check), decreasing = TRUE)
```

::: callout-note
## What "data leakage" means and how we caught it

- **Data leakage** occurs when information from the outcome (or its
  future) is used to *build* the predictors. The model then appears
  brilliant but has learned nothing — it is just recovering an identity.
- Step 1 spotted the tell-tale signs: identical coefficients and
  implausibly large t-values.
- Step 2 confirmed it: `cor()` returned exactly 1, and the ratio
  `incidence / (female + male)` was a constant for every country.
- Step 3 generalised the search. `st_drop_geometry()` removes the
  spatial column (which is not numeric); `select(where(is.numeric))`
  keeps only numeric columns; `select(-incidence)` drops the outcome;
  then `sapply()` computes the correlation of each remaining column with
  `incidence`. Sorting by absolute value surfaces every leaked variable
  at once.
:::

The output shows the leakage extends well beyond `female`/`male`. Almost
every variable with `|r| > 0.85` shares a consistent naming pattern:
age-banded (`age*`), year-indexed (`yr*`) and male/female-stratified
(`m*`/`f*`). This strongly indicates they are all components of the same
underlying case-count calculation that produced `incidence`.

------------------------------------------------------------------------

## Refined Model {#sec-ols-refine}

Based on the correlation screen, we exclude every variable structurally
tied to the incidence calculation along with the non-informative row
identifier. In their place we retain our six simulated risk-factor
covariates.

Another thing worth mentioning in epidemiology is the raw incidence
counts. We rarely run models on raw count data. Instead we will
calculate an **incidence rate** (e.g. cases per 100,000 people) before
modelling.

```{r refined-ols}
africa_health <- africa_health |> 
  mutate(incidence_rate = (incidence / TOTPOP) * 100000)

ols_refined <- lm(
  incidence_rate ~ smoking_prev + obesity_prev + diabetes_prev + 
    alcohol_per_capita + family_history_prev + urbanization_pct,
  data = africa_health
)
summary(ols_refined)
```

::: callout-note
## Rates vs counts

Raw case counts are dominated by population size: a big country has more
cases almost by definition. Dividing by population —
`(incidence / TOTPOP) * 100000` — standardises the outcome to "cases per
100,000 people", making countries comparable. This is the same reasoning
behind the age-standardised rates used in published cancer statistics.

With the leaked columns removed, this refined model has an $R^2$ near
0.15 and no significant predictors. Because the covariates were pure
random noise, OLS did exactly what it should: it found **no** real
relationship. The two variables that show marginal significance
(`obesity_prev`, `urbanization_pct`) are chance findings — their signs
are even biologically backwards (negative), which is what random noise
looks like.
:::

This output provides an overall statistically insignificant model.
Because the simulated covariates were generated purely as random noise
(`runif`), the OLS model did exactly what it is supposed to do. It shows
no statistical relationship between these covariates and the target
variable.

Although all the covariates are randomly generated, `obesity_prev` and
`urbanization_pct` are showing marginal significance. This is purely by
chance, because we can confirm from their estimate direction (`-0.103`
and `-0.041`) that the math contradicts the biology; higher obesity and
urbanisation should increase pancreatic cancer risk.

------------------------------------------------------------------------

## Test for Spatial Residual Autocorrelation {#sec-residual-moran}

**Purpose:** Check whether the OLS model's residuals are spatially
random. The core assumption of OLS is that its errors (residuals) are
randomly distributed. If significant autocorrelation remains, OLS
standard errors are biased and a spatial model (or GWR) is needed.

In spatial epidemiology, if our model is missing a key geographical
variable, the model's errors will cluster together geographically. This
is called **spatial autocorrelation of residuals**.

```{r moran-residuals}
residual_moran <- lm.morantest(
  model = ols_refined,
  listw = lw,
  zero.policy = TRUE
)
print(residual_moran)
```

::: callout-note
## The zero-neighbour warning, and what `zero.policy = TRUE` does

Six of the 54 units are islands that share no border with any other unit
in the layer: `Cabo Verde`, `Comoros`, `Sao Tome and Principe`,
`Madagascar`, `Mauritius` and `Seychelles`. `poly2nb()` warns about this
and reports 7 sub-graphs — the connected mainland block of 48, plus the
six islands. The warning is expected: these are genuine islands

A unit with no neighbours has an undefined spatial lag, so `spdep` has
to be told what to do with those rows. `zero.policy = TRUE` keeps them
in the analysis and sets their lag to `0`. Setting it to `FALSE` fails
instead, and not always in the same way: `nb2listw()` stops with an
`Empty neighbour sets found` error, `moran.test()` stops with a
`regions with no neighbours found` error, and `lm.morantest()`
quietly returns `NA`. That is why the weights object and every test
in this module pass `zero.policy = TRUE`.

The islands still enter the statistic through their own residuals, but
they appear in no other unit's neighbour list. They add nothing to the
numerator of Moran's I, yet still count in the denominator (the sum of
squared deviations), so keeping them in pulls I slightly towards zero.
:::

::: callout-note
## Why test the residuals, not the outcome

The **residual** for each country is the part of its rate the model
*could not* explain: `residual = observed − fitted`. If those leftovers
are spatially clustered (nearby countries share similar errors), the
model is missing a spatially structured cause, and its p-values are
unreliable.

`lm.morantest()` applies Moran's I (from Module 3) to the residuals
instead of the raw outcome. The null hypothesis is that the residuals
are randomly distributed across space.
:::

**Key parameters:** - `lm.morantest()`: Applies Moran's I to the
regression residuals rather than the raw outcome.

**Expected output:** - If p \< 0.05, residuals are spatially
autocorrelated, so OLS is likely misspecified and a spatial model (like
GWR) is worth trying — but whether it actually helps is settled by
comparing AICc (see below), not by the test alone. - If p \> 0.05, the
residuals are not *significantly* autocorrelated — but a p-value close
to 0.05 (say 0.06) signals weak, inconclusive spatial structure, not
genuinely random residuals.

Here the observed Moran's I of the residuals is **+0.126** against an
expectation of **−0.024**, with a p-value of **0.065**. That p-value is
one-sided (`alternative hypothesis: greater`): it tests whether the
residuals are *more* clustered than chance, not whether they differ from
chance in either direction.

At the 5% level we do **not** reject the null hypothesis, so the
residuals are not *significantly* spatially autocorrelated. But 0.065 is
a borderline result rather than a comfortable one: the observed I is
positive — nearby countries do share similar errors — and it misses the
conventional threshold only narrowly. The defensible reading is “weak,
inconclusive spatial structure in the residuals”, not “residuals are
spatially random”.

------------------------------------------------------------------------

## Geographically Weighted Regression (GWR) {#sec-gwr}

**Purpose:** Fit local regression models that allow the relationship
between cancer and lifestyle/genetic risk factors to vary across space
(spatial non-stationarity).

Ordinary regression fits **one** equation for the whole continent. GWR
fits a separate weighted regression at every location, giving nearby
points more influence, so each country gets its own set of coefficients.

```{r gwr-model}
# Step 1: Find optimal adaptive bandwidth (number of nearest neighbours)
# This uses cross-validation to find best fit
bw_adapt <- bw.gwr(
  formula = incidence_rate ~ smoking_prev + obesity_prev + 
    diabetes_prev + alcohol_per_capita + 
    family_history_prev + urbanization_pct,
  data = africa_health,
  approach = "CV",     # Cross-validation approach
  kernel = "bisquare", # the shape of the spatial weight decay
  adaptive = TRUE      # TRUE = find optimal no. of neighbours. FALSE = find distance
)
cat("Optimal adaptive bandwidth:", bw_adapt, "neighbors\n")

# Step 2: Fit GWR model
gwr_model <- gwr.basic(
  formula = incidence_rate ~ smoking_prev + obesity_prev + 
    diabetes_prev + alcohol_per_capita + 
    family_history_prev + urbanization_pct,
  data = africa_health,
  bw = bw_adapt,
  kernel = "bisquare",
  adaptive = TRUE
)

# Print summary (includes global OLS + local GWR diagnostics)
print(gwr_model)
```

::: callout-note
## How GWR works, step by step

- **Bandwidth** is the size of the local "window". `bw.gwr()` searches
  over bandwidths and picks the one that minimises cross-validation
  error (`approach = "CV"`). With `adaptive = TRUE`, the bandwidth is a
  *number of nearest neighbours* (35 here) rather than a distance — the
  right choice for irregularly sized countries.
- **Kernel** is the weighting function. `"bisquare"` gives full weight
  to near neighbours and drops smoothly to zero at the edge of the
  window, creating clean local windows. (`"gaussian"` is the other
  common choice; it never quite reaches zero.)
- `gwr.basic()` then fits a weighted OLS at every polygon centroid. The
  result is one coefficient per predictor **per country** — hence the
  coefficients are reported as ranges.
:::

**Key parameters:** - `bw.gwr()`: `adaptive = TRUE` finds a bandwidth in
number of neighbours rather than distance units. Better for irregularly
sized polygons. - `kernel = "bisquare"`: Weights drop to zero beyond the
bandwidth, creating clean local windows. - `gwr.basic()`: Fits a
separate weighted OLS at every polygon centroid.

**Expected output:** - Bandwidth value (35 neighbours for country-level
Africa data). - GWR diagnostics showing local R², coefficient ranges,
and AICc values vs global OLS.

::: callout-note
## Judging whether GWR is worth it

A bandwidth of 35 is huge for a study area of only 54 countries — the
model must look at over half the continent to make a single local
prediction. The coefficient range for `diabetes_prev` spans roughly
−0.64 to +0.23, suggesting the relationship *appears* to flip sign in
places. But this is simulated noise, so the model is chasing nothing.

The honest way to judge GWR is **AICc** (lower is better), not $R^2$ —
GWR almost always inflates $R^2$ because it fits many more parameters.
Here the AICc *rose* from 276.4 for OLS to 294.7 for GWR. The local
model does not earn back the parameters it spends, so on this evidence
it is the worse model — a useful negative result.

The residual Moran test is only weak corroboration here (I = 0.126,
p = 0.065), and its verdict is specific to this covariate set — see
below.
:::

A bandwidth of 35 is huge for a spatial region of 54 countries. The
model is forced to look at over half a continent to make a single
prediction. The GWR coefficient estimates are given as a range because
they change depending on where you are on the map. In the global OLS the
relationship for `diabetes_prev` is negative, but GWR reveals that it
changes in some part of the continent (min ≈ −0.64 and max ≈ 0.23). This
shows the relationship is unstable across space (but remember this is
simulated data, so the model is just chasing random noise).

The GWR $R^2$ looks very high but that should not be the metric of
interest, because it is almost always high for GWR. The parameter of
interest is the AICc (lower is better). Even though the GWR artificially
inflated the $R^2$, the AICc went up from 276.4 in the OLS to 294.7 —
the GWR is the worse model once you account for the number of parameters
it spends.

Keep the scope of that verdict in mind. It is a statement about *this*
six-covariate specification, not a general finding that spatial
regression is unnecessary. Two things narrow it. First, the covariates
were simulated at random, so there is no spatially varying relationship
for GWR to recover — the local model can only chase noise. Second, the
residual evidence is borderline rather than clean: the global model
leaves I = 0.126 (p = 0.065), which fails to reach significance but is
not the comfortable “no clustering at all” result that would make GWR
moot. Change the covariate set and the balance can shift the other way —
the two-covariate model in the exercise below leaves much stronger
residual clustering (I = 0.217, p = 0.007), which is precisely the
situation GWR is designed for. The practical lesson is to let the AICc
decide for each specification instead of assuming that a residual test
has ruled spatial models out in advance.

------------------------------------------------------------------------

## Interpret and Visualize GWR Results {#sec-gwr-viz}

**Purpose:** Map local coefficient surfaces to see **where** each
predictor matters most.

```{r gwr-maps}
# gwr_model$SDF is an sf object containing local coefficients
gwr_sf <- gwr_model$SDF

# Map the spatially varying coefficient for diabetes
tm_shape(gwr_sf) +
  tm_polygons(
    fill = "diabetes_prev",
    fill.scale = tm_scale_continuous(
      values = "-brewer.rd_bu",     # Negative RdBu puts red at positive and blue at negative
      midpoint = 0                  # Forces the white/neutral colour to sit exactly at zero
    ),
    fill.legend = tm_legend(title = "Local Coefficient\n(Diabetes)"),
    col = 'black',
    col_alpha = 0.5,        
    lwd = 0.5               
  ) +
  tm_title("GWR: Spatially Varying Effect of Diabetes")

# Map the local model fit (Local R-squared)
tm_shape(gwr_sf) +
  tm_polygons(
    fill = "Local_R2",
    fill.scale = tm_scale_continuous(
      values = "viridis"
    ),
    fill.legend = tm_legend(title = "Local R-Squared"),
    col = "black",
    lwd = 0.5
  ) +
  tm_title("GWR: Local Model Performance across Africa")
```

::: callout-note
## Reading the coefficient maps

- `gwr_model$SDF` is the spatial data frame of local results, with one
  row per country and a column for each local coefficient plus
  `Local_R2`.
- `values = "-brewer.rd_bu"` selects a red–blue *diverging* palette (the
  `-` reverses it so red = positive). `midpoint = 0` pins the neutral
  colour at zero, so red countries have a positive local coefficient and
  blue ones a negative one — the sign is instantly readable.
- The second map shows `Local_R2`, the model's fit at each location,
  using a sequential viridis palette.
:::

**Key parameters:** - `midpoint = 0` in
`fill.scale = tm_scale_continuous()` centers the colour scale at zero,
making positive (red) and negative (blue) effects immediately visible.

**Expected output:** Two maps; one showing the spatially varying
coefficient `diabetes_prev` and the other showing where the GWR model
performs best (`Local_R2`). Red indicates a positive local association;
blue indicates a negative local association for `diabetes_prev`.

The spatial variation map shows smooth colour transitions across
borders. While this creates a beautiful map, remember our **AICc**
score. Because the underlying data is purely random noise, this map is
just a visualisation of the GWR algorithm "forcing" a spatial pattern
where none actually exists.

------------------------------------------------------------------------

## Practical Exercise {#sec-m4-exercise}

Reinforce the workflow with a few tasks, then check your answers against
the solution.

1.  Confirm the data leakage: compute
    `cor(africa_health$incidence, africa_health$female + africa_health$male)`
    and report the value.
2.  Create the `incidence_rate` variable and fit an OLS model using only
    `smoking_prev` and `obesity_prev`. Report the $R^2$.
3.  Run `lm.morantest()` on that model's residuals. Is there significant
    spatial autocorrelation?
4.  *(Challenge)* Re-run `bw.gwr()` with `kernel = "gaussian"` (keeping
    the other settings) and report the optimal bandwidth and whether the
    GWR AICc improved relative to the `"bisquare"` model above.

::: {.callout-note collapse="true"}
## Model solution

``` r
library(sf)
library(spdep)
library(GWmodel)
library(dplyr)
library(mlspatial)

data(africa_shp, package = "mlspatial")
data(panc_incidence, package = "mlspatial")
africa_health <- join_data(
  africa_shp |> mutate(NAME = gsub("Côte d'Ivoire", "Cote d'Ivoire", NAME)),
  panc_incidence, by = "NAME"
)
africa_health <- st_make_valid(africa_health)

nb <- poly2nb(africa_health, row.names = africa_health$NAME,
              queen = TRUE, snap = 1000)
lw <- nb2listw(nb, style = "W", zero.policy = TRUE)

set.seed(898)
n <- nrow(africa_health)
africa_health <- africa_health |>
  mutate(
    smoking_prev = round(runif(n, 8, 28), 1),
    obesity_prev = round(runif(n, 5, 30), 1),
    diabetes_prev = round(runif(n, 2, 11), 1),
    alcohol_per_capita = round(runif(n, 0.5, 12), 1),
    family_history_prev = round(runif(n, 3, 10), 1),
    urbanization_pct = round(runif(n, 15, 85), 1),
    incidence_rate = (incidence / TOTPOP) * 100000
  )

# 1. Confirm the leakage
cor(africa_health$incidence, africa_health$female + africa_health$male)
# -> exactly 1

# 2. OLS with two covariates
ols2 <- lm(incidence_rate ~ smoking_prev + obesity_prev, data = africa_health)
summary(ols2)$r.squared

# 3. Residual autocorrelation
lm.morantest(ols2, listw = lw, zero.policy = TRUE)
# -> I = 0.217, p = 0.007: with only two covariates the residuals ARE
#    significantly clustered - a weakly fitting global model leaves
#    spatial structure behind

# 4. Gaussian kernel bandwidth + AICc
f <- incidence_rate ~ smoking_prev + obesity_prev + diabetes_prev +
  alcohol_per_capita + family_history_prev + urbanization_pct
bw_gauss <- bw.gwr(f, data = africa_health, approach = "CV",
                   kernel = "gaussian", adaptive = TRUE)
gwr_gauss <- gwr.basic(f, data = africa_health, bw = bw_gauss,
                       kernel = "gaussian", adaptive = TRUE)
gwr_gauss$GW.diagnostic$AICc
# -> bandwidth 24, AICc = 279.7: this beats the bisquare GWR (294.7)
#    but is still above the six-covariate OLS AICc of 276.4, so the
#    Gaussian kernel narrows the gap without changing the verdict
```
:::

------------------------------------------------------------------------

## Module 4 Checklist {#sec-m4-check .unnumbered}

- [ ] Simulated risk-factor covariates
- [ ] Built and interpreted an OLS regression model
- [ ] Diagnosed data leakage from a perfect fit
- [ ] Tested OLS residuals for spatial autocorrelation
- [ ] Fitted a GWR model with adaptive bandwidth
- [ ] Mapped local coefficient surfaces
- [ ] Practical exercise completed

**Next:** Proceed to [Module 5: Machine Learning for Spatial
Epidemiology](05-ml.qmd).
