---
editor: 
  markdown: 
    wrap: 72
---

# Module 4: Spatial Regression and Epidemiological Risk Factor Analysis {#sec-module4}

This module connects health outcomes to lifestyle/genetic drivers. You
will simulate covariates, build a global OLS model, test its residuals
for spatial autocorrelation, and then fit a Geographically Weighted
Regression (GWR) to capture spatially varying relationships.

```{r setup-m4}
#| include: false
library(sf)
library(terra)
library(spdep)
library(GWmodel)
library(dplyr)
library(mlspatial)
library(tmap)

data(africa_shp, package = "mlspatial")
data(panc_incidence, package = "mlspatial")
africa_health <- join_data(
  africa_shp |> mutate(
    across(c(NAME, COUNTRYAFF),
           ~ replace_values(.x, "Côte d'Ivoire" ~ "Cote d'Ivoire"))),
  panc_incidence, by = "NAME"
  )
africa_health <- st_make_valid(africa_health)

nb <- poly2nb(africa_health, row.names = africa_health$NAME, queen = TRUE)
lw <- nb2listw(nb, style = "W", zero.policy = TRUE)
```

------------------------------------------------------------------------

## Simulate Risk-Factor Covariates {#sec-simulate}

**Purpose:** Link health outcomes to known epidemiological risk factors
for pancreatic cancer (smoking, obesity, diabetes, alcohol use, family
history urbanisation) for use in spatial regression exercise

For this tutorial, we **simulate** six covariates directly at the
country level so the code is fully runnable without downloading external
datasets. Because `africa_health` is a polygon layer where each row
already represents one country (like `incidence`, `TOTPOP` etc), we do
not need a raster data - we can generate one random value per row and
append it directly with `mutate()`

**In real research**, replace these random draws with actual
country-level estimates from sources such as the WHO Global Health
Observatory (smoking, obesity, diabetes prevalence), IARC GLOBOCAN
risk-factor tables, or national health surveys. These are typically
small CSV/Excel downloads, joined to `africa_health` by country ISO code
(`ISO_3DIGIT`) rather than simulated.

```{r simulate-covariates}
set.seed(898)  # reproducibility

n <- nrow(africa_health)

africa_health <- africa_health |> 
  mutate(
    # Smoking prevalence (% of adult population) — WHO African region estimates
    # typically range ~10-25%; strongest modifiable risk factor for pancreatic cancer
    smoking_prev = round(runif(n, min = 8, max = 28), 1),
    
    # Obesity prevalence (% adults, BMI >= 30) — WHO Africa region estimates
    # range roughly 5-30% depending on urbanization/country
    obesity_prev = round(runif(n, min = 5, max = 30), 1),
    
    # Type 2 diabetes prevalence (%) — longstanding diabetes is a well-established
    # pancreatic cancer risk factor; IDF Africa estimates ~3-10% typically
    diabetes_prev = round(runif(n, min = 2, max = 11), 1),
    
    # Alcohol consumption (litres pure alcohol per capita per year) —
    # heavy long-term use linked to chronic pancreatitis -> elevated risk
    alcohol_per_capita = round(runif(n, min = 0.5, max = 12), 1),
    
    # Family history / genetic predisposition proxy (% of cases with
    # first-degree relative history) — literature suggests ~5-10% of
    # pancreatic cancers have a hereditary component
    family_history_prev = round(runif(n, min = 3, max = 10), 1),
    
    # Urbanization (%) — proxy for dietary/lifestyle transition,
    # often used as a covariate for NCD risk in African epidemiology
    urbanization_pct = round(runif(n, min = 15, max = 85), 1)
  )

head(africa_health[c("NAME", "smoking_prev", "obesity_prev", 
                     "diabetes_prev", "alcohol_per_capita", 
                     "family_history_prev", "urbanization_pct")])
```

**Key parameters:** - `runif(n, min, max)`: draws `n` random values from
a uniform distribution bounded to a realistic range for each covariate,
loosely based on published prevalence ranges across African countries.

These covariates are simulated **independently of `incidence` and of
each other** — they are not designed to correlate with pancreatic cancer
rates. This is intentional: the goal is to demonstrate the spatial
regression *pipeline* (data structure, model syntax, diagnostics), not
to produce a real epidemiological finding.

**Expected output:** Six new columns appended to `africa_health`

| Column | Description | Simulated range |
|------------------------|------------------------|------------------------|
| `smoking_prev` | Adult smoking prevalence (%) | 8–28% |
| `obesity_prev` | Adult obesity prevalence, BMI ≥ 30 (%) | 5–30% |
| `diabetes_prev` | Type 2 diabetes prevalence (%) | 2–11% |
| `alcohol_per_capita` | Alcohol consumption (litres pure alcohol/year) | 0.5–12 |
| `family_history_prev` | Cases with first-degree relative history (%) | 3–10% |
| `urbanization_pct` | Urban population (%) | 15–85% |

------------------------------------------------------------------------

## Build Ordinary Least Squares (OLS) Regression {#sec-ols}

**Purpose:** Establish a global (non-spatial) baseline model relating
pancreatic cancer incidence to lifestyle, genetic and demographic risk
factors.

```{r ols-model}
ols_model <- lm(
  incidence ~ smoking_prev + obesity_prev + diabetes_prev + 
    alcohol_per_capita + family_history_prev + urbanization_pct + 
    female + male + ageb + agec,
  data = africa_health
)
summary(ols_model)
```

**Key parameters:** - Formula syntax:
`response ~ predictor1 + predictor2 + ...` - Interpretation: Positive
coefficients indicate higher expected incidence with higher predictor
values.

**Expected output:** A standard `lm` summary showing coefficients,
standard errors, t-values, and p-values. R² indicates the proportion of
variance explained globally.

This model prints a warning message. The $R^2$ value is exactly **1**
which creates room for concern. This is not a real spatial regression
breakthrough, it is a classic case of data leakage through an identity
relationship. Hence the perfect fit is a red flag not a good sign.

------------------------------------------------------------------------

## Diagnosing Data Leakage {#sec-diagnose}

The first **ols** model returned `R-squared: 1`, an F-statistic of
`2.36e+31` and an explicit warning -
`essentialperfect fit: summary may be unreliable`. A model that fits
perfectly is almost always telling you that the outcome is
mathematically derived from one or more of its own predictors. This is
called **data leakage**

**Step 1: Inspect the coefficients.** Looking at the model summary, two
variables stood out immediately:

```         
female   9.302e-02   t = 1.005e+13   p < 2e-16
male     9.302e-02   t = 1.003e+13   p < 2e-16
```

Both coefficients are identical to four decimal figures, with t-value
around $10^{13}$. T-values that enormous are a signature of the model
detecting an algebraic identity not a strong effect.

**Step 2: Test the suspected Identity directly**

```{r test-suspect}
cor(africa_health$incidence, africa_health$female + africa_health$male)
head(africa_health$incidence / (africa_health$female + africa_health$male))
```

A correlation of 1 between `incidence` and gender confirms that
`incidence` is a fixed linear function of `female + male`. Dividing one
by the other reveals the exact relationship. Every single country
returns the exact constant `0.09302326`, meaning `incidence` was
computed directly from `male` and `female` (or they share a common
upstream calculation)

Including `male` and `female` as predictors makes it **mathematically
circular**, the model is not learning but rather reversing an arithmetic
formula.

**Step 3: Screen every numeric variable for leakage**

```{r check-cor}
# Check all numeric columns systematically against incidence
numeric_cols <- africa_health |> 
  st_drop_geometry() |>                    # drop geometry — not numeric
  select(where(is.numeric)) |> 
  select(-incidence) |>                    # exclude the outcome itself
  names()

cor_check <- sapply(numeric_cols, function(col) {
  cor(africa_health$incidence, africa_health[[col]], use = "complete.obs")
})

sort(abs(cor_check), decreasing = TRUE)

```

The output shows the leakage extends well beyond `female`/`male`. Almost
every variable with `|r|>0.85` shares a consistent naming pattern:
age-banded (`age*`), year-indexed (`yr*`) and male/female-stratified
(`m*`/`f*`). This strongly indicates they are all components of the same
underlying case-count calculation that produced `incidence`.

------------------------------------------------------------------------

## Refined Model {#sec-ols-refine}

Based on the correlation screen, we exclude every variable structurally
tied to incidence calculation along with the non-informative row
identifier. In their place we retain our six simulated risk-factor
covariates.

Another thing worth mentioning in epidemiology is the raw incidence
counts. We rarely run models on raw count data. Instead we will
calculate an **incidence rate** (e.g cases per 100,000 people) before
modelling.

```{r refined-ols}
africa_health <- africa_health |> 
  mutate(incidence_rate = (incidence/TOTPOP)*100000)

ols_refined <- lm(
  incidence_rate ~ smoking_prev + obesity_prev + diabetes_prev + 
    alcohol_per_capita + family_history_prev + urbanization_pct,
  data = africa_health
)
summary(ols_refined)
```

This output provides an overall statistically insignificant model.
Because the simulated covariates were generated purely as random noise
(`runif`), the OLS model did exactly what it is supposed to do. It shows
no statistical relationship between these covariates and the target
variable.

Although all the covariates are randomly generated, `obesity_prev` and
`urbanization_pct` are showing marginal significance. This is purely by
chance because we can confirm from their estimate direction (`-0.103`
and `-0.041`) that the math contradicts the biology; higher obesity and
urbanisation should increase pancreatic cancer risk.

------------------------------------------------------------------------

## Test for Spatial Residual Autocorrelation {#sec-residual-moran}

**Purpose:** Check whether the OLS model's residuals are spatially
random. The core assumption of OLS is that its errors (residuals) are
randomly distributed. If significant autocorrelation remains, OLS
standard errors are biased and a spatial model (or GWR) is needed.

In spatial epidemiology, if our model is missing a key geographical
variable, the model's error will cluster together geographically. This
is called **Spatial Autocorrelation of Residuals**

```{r moran-residuals}
residual_moran <- lm.morantest(
  model = ols_refined,
  listw = lw,
  zero.policy = TRUE
)
print(residual_moran)
```

**Key parameters:** - `lm.morantest()`: Applies Moran's I to the
regression residuals rather than the raw outcome.

**Expected output:** - If p \< 0.05, residuals are spatially
autocorrelated → OLS is misspecified, therefore a spatial model (like
GWR) is required. - If p \> 0.05, OLS residuals are spatially random
(rare in epidemiology).

The Null hypothesis for this test is that the residuals are randomly
distributed (no spatial autocorrelation). Because our p-vale is high, we
conclude that the errors in the model are randomly distributed across
the African continent. There is no statistically significant spatial
autocorrelation in the residuals.

------------------------------------------------------------------------

## Geographically Weighted Regression (GWR) {#sec-gwr}

**Purpose:** Fit local regression models that allow the relationship
between cancer and lifestyle/genetic risk factors to vary across space
(spatial non-stationarity).

```{r gwr-model}
# Step 1: Find optimal adaptive bandwidth (number of nearest neighbours)
# This uses cross-validation to find best fit
bw_adapt <- bw.gwr(
  formula = incidence_rate ~ smoking_prev + obesity_prev + 
    diabetes_prev + alcohol_per_capita + 
    family_history_prev + urbanization_pct,
  data = africa_health,
  approach = "CV",     # Cross-validation approach
  kernel = "bisquare", # the shape of the spatial weight decay
  adaptive = TRUE      # TRUE=find optimal no. of neigbours. FALSE = find distance
)
cat("Optimal adaptive bandwidth:", bw_adapt, "neighbors\n")

# Step 2: Fit GWR model
gwr_model <- gwr.basic(
  formula = incidence_rate ~ smoking_prev + obesity_prev + 
    diabetes_prev + alcohol_per_capita + 
    family_history_prev + urbanization_pct,
  data = africa_health,
  bw = bw_adapt,
  kernel = "bisquare",
  adaptive = TRUE,
)

# Print summary (includes global OLS + local GWR diagnostics)
print(gwr_model)
```

**Key parameters:** - `bw.gwr()`: `adaptive = TRUE` finds a bandwidth in
number of neighbours rather than distance units. Better for irregularly
sized polygons. - `kernel = "bisquare"`: Weights drop to zero beyond the
bandwidth, creating clean local windows. - `gwr.basic()`: Fits a
separate weighted OLS at every polygon centroid.

**Expected output:** - Bandwidth value (gave 35 neighbours for
country-level Africa data). - GWR diagnostics showing local R²,
coefficient ranges, and AICc values vs global OLS.

A bandwidth of 35 is huge for a spatial region of 54 countries. The
model is forced to look at over half a continent to make a single
prediction. The GWR coefficient estimates are given as a range because
it changes depending on where you are on the map. In the global OLS the
relationship for `diabetes_pev` is negative but GWR reveals that it
changes in some part of the continent (`min:-0.543` and `max:0.225`).
This shows the relationship is unstable across space (But remember this
is simulated data so the model is just chasing random noise).

The GWR $R^2$ looks very high but that should not be the metric of
interest because it is almost always high for GWR analysis. The
parameter of interest is the AICc (Lower is better). Even the the GWR
artificially inflated the $R^2$, the AICc actually went up from 276.4 in
the OLS to 294.7, this tells us that the GWR model is statistically
worse. The OLS errors were randomly distributed and so the GWR was
completely unnecessary.

------------------------------------------------------------------------

## Interpret and Visualize GWR Results {#sec-gwr-viz}

**Purpose:** Map local coefficient surfaces to see **where** each
predictor matters most.

```{r gwr-maps}
# gwr_model$SDF is an sf object containing local coefficients
gwr_sf <- gwr_model$SDF

# Map the spatially varying coefficient for diabetes
tm_shape(gwr_sf) +
  tm_polygons(
    fill = "diabetes_prev",
    fill.scale = tm_scale_continuous(
      values = "-brewer.rd_bu",     # Negative RdBu puts red at positive and Blue at negative
      midpoint = 0          # Forces the white/neutral color to sit exactly at zero
    ),
    fill.legend = tm_legend(title = "Local Coefficient\n(Diabetes)"),
    col = 'black',
    col_alpha = 0.5,        
    lwd = 0.5               
  ) +
  tm_title("GWR: Spatially Varying Effect of Diabetes")

# Map the local model fit (Local R-squared)
tm_shape(gwr_sf) +
  tm_polygons(
    fill = "Local_R2",
    fill.scale = tm_scale_continuous(
      values = "viridis"
    ),
    fill.legend = tm_legend(title = "Local R-Squared"),
    col = "black",
    lwd = 0.5
  ) +
  tm_title("GWR: Local Model Performance across Africa")

```

**Key parameters:** - `midpoint = 0` in
`fill.scale = tm_scale_continuous()` centers the colour scale at zero,
making positive (red) and negative (blue) effects immediately visible.

**Expected output:** Two maps; one showing spatially varying coefficient
`diabetes_prev` and the other showing the the area where the GWR model
performs best (`Local R2`). Areas with red indicate a positive local
association; blue indicates a negative local association for the
`diabetes_prev` .

The spatial variation map shows smooth colour transitions across
borders. Whiles this creates a beautiful map, remember our **AICc**
score. Because the underlying data is purely random noise, this map is
just a visualisation of the GWR algorithm 'forcing' a spatial pattern
where none actually exists.

------------------------------------------------------------------------

## Module 4 Checklist {#sec-m4-check .unnumbered}

- [ ] Extracted environmental covariates from raster data
- [ ] Built and interpreted an OLS regression model
- [ ] Tested OLS residuals for spatial autocorrelation
- [ ] Fitted a GWR model with adaptive bandwidth
- [ ] Mapped local coefficient surfaces

**Next:** Proceed to [Module 5: Machine Learning for Spatial
Epidemiology](05-ml.qmd).
