---
editor: 
  markdown: 
    wrap: sentence
---

# Module 5: Machine Learning for Spatial Epidemiology (Showcasing `mlspatial`) {#sec-module5}

This module showcases the machine learning capabilities of `mlspatial`.
You will prepare data, train three algorithms (Random Forest, XGBoost, Support Vector Regression), evaluate their performance, and map predictions to identify spatial risk patterns.

**What you will learn**

- How to turn an `sf` object into a clean table ready for ML.
- What Random Forest, XGBoost, and Support Vector Regression are, in plain terms.
- The difference between **in-sample** fit and **held-out** (cross-validated) performance — and why it matters.
- How to run honest, cross-validated evaluation with `caret` and `xgboost`.
- How to map model predictions back onto the map.

```{r setup-m5}
#| include: false
library(sf)
library(terra)
library(dplyr)
library(tidyr)
library(mlspatial)
library(tmap)
library(randomForest)
library(xgboost)
library(e1071)
library(caret)

data(africa_shp, package = "mlspatial")
data(panc_incidence, package = "mlspatial")
africa_health <- join_data(
  africa_shp |> mutate(
    NAME       = gsub("Côte d'Ivoire", "Cote d'Ivoire", NAME),
    COUNTRYAFF = gsub("Côte d'Ivoire", "Cote d'Ivoire", COUNTRYAFF)
  ),
  panc_incidence, by = "NAME"
)
africa_health <- st_make_valid(africa_health)

# Add simulated environmental data
set.seed(898)  
n <- nrow(africa_health)
africa_health <- africa_health |> 
  mutate(
    incidence_rate = (incidence/TOTPOP)*100000,
    smoking_prev = round(runif(n, min = 8, max = 28), 1),
    obesity_prev = round(runif(n, min = 5, max = 30), 1),
    diabetes_prev = round(runif(n, min = 2,  max = 11), 1),
    alcohol_per_capita = round(runif(n, min = 0.5, max = 12), 1),
    family_history_prev = round(runif(n, min = 3,  max = 10), 1),
    urbanization_pct = round(runif(n, min = 15, max = 85), 1)
  )
```

------------------------------------------------------------------------

## Introduction to `mlspatial` for ML {#sec-ml-intro}

**Purpose:** Understand what `mlspatial` provides and how it fits into a spatial epidemiology ML workflow, before writing any modelling code.

Traditional spatial epidemiology often requires switching between GIS software, R scripts, and Python notebooks.
`mlspatial` (Azeez & Noel, 2025) packages together spatial data handling (`sf`), thematic mapping (`tmap`), spatial autocorrelation diagnostics (`spdep`), and machine learning (`randomForest`, `xgboost` and `e1071`) into a single applied workflow aimed at spatial epidemiologists.

The package provides a number of convenient functions and leans on well-established packages for this workflow.
In this module we will train three ML algorithms — Random Forest, XGBoost, and Support Vector Regression — to predict pancreatic cancer incidence from our simulated covariates.

::: callout-note
## What each `mlspatial` training function does

The `train_*()` functions are thin, friendly wrappers over the underlying
packages — useful to know because the output is exactly the same as if
you had called the underlying package yourself:

- `train_rf(data, formula, ntree, seed)` calls
  `randomForest::randomForest(formula, data, ntree, importance = TRUE)`
  after fixing the seed.
- `train_xgb(data, formula, nrounds, max_depth, learning_rate)` builds
  the model matrix and label internally, then runs
  `xgboost::xgb.train()` with a squared-error objective.
- `train_svr(data, formula)` calls `e1071::svm()` with
  `type = "eps-regression"` and a radial kernel.

The rest of this module explains *what those choices mean* and, most
importantly, how to evaluate the results honestly.
:::

------------------------------------------------------------------------

## Prepare Data for ML Modelling {#sec-ml-prep}

**Purpose:** Prepare our `sf` object into a clean, geometry-free data frame for machine learning algorithms.
We drop the geometry column, select only relevant columns and check for missing values, and decide whether scaling is needed for each algorithm.

```{r ml-prep}
# Drop geometry and select complete cases
ml_data <- africa_health |> 
  st_drop_geometry() |> 
  select(
    incidence_rate,                    # outcome
    smoking_prev, obesity_prev, diabetes_prev,
    alcohol_per_capita, family_history_prev,
    urbanization_pct, TOTPOP                 
  ) |> 
  drop_na()

# Quick structural check
str(ml_data)
sum(is.na(ml_data))  
```

::: callout-note
## Why this step exists

Most ML algorithms expect a plain rectangular table of numbers — no
geometry column, no missing values. So we:

- `st_drop_geometry()` — removes the spatial column. The spatial link is
  not lost: the *row order* of `ml_data` still matches `africa_health`, so
  we can paste predictions back onto the map later (as we do in
  @sec-patterns).
- `select(...)` — keeps only the outcome and predictors we actually want.
- `drop_na()` — removes rows with missing values. XGBoost and SVR will
  error on `NA`, so this is not optional.
:::

**Key parameters:** - `st_drop_geometry()`: ML algorithms cannot use sf geometry columns; we keep the spatial link via row order for later mapping.
- `drop_na()`: XGBoost and SVR do not tolerate missing values.

**Expected output:** A plain `data.frame` (no `sf`/`data.frame` dual class) with remaining rows after removing `NA`s, and 8 columns.

::: callout-note
## On scaling

**Tree-based methods** (Random Forest, XGBoost) are **scale-invariant**:
they make decisions by comparing one value against a threshold, so
multiplying a predictor by a constant changes nothing. Standardising
predictors therefore has no effect on their fit.

**SVR**, in contrast, measures similarity using Euclidean distance in
feature space, so a variable with a large range (like `TOTPOP`, in the
millions) would dominate one with a small range (like a percentage).
That is why we scale the predictors — but only immediately before the SVR
fit in @sec-svr, leaving `ml_data` itself unscaled for the other two
models.
:::

------------------------------------------------------------------------

## Train a Random Forest Model {#sec-rf}

**Purpose:** Use an ensemble of decision trees to capture non-linear relationships between predictors and `incidence_rate`, and inspect variable importance.

```{r train-rf}
# Train Random Forest regression model
rf_formula <- incidence_rate ~ smoking_prev + obesity_prev + diabetes_prev +
  alcohol_per_capita + family_history_prev + urbanization_pct + TOTPOP

rf_model <- mlspatial::train_rf(
  data = ml_data,
  formula = rf_formula,
  ntree = 500,
  seed = 222
)

print(rf_model)
```

::: callout-note
## What Random Forest is

A **decision tree** repeatedly splits the data on a predictor (e.g.
"smoking > 20%?") to form groups with similar outcomes. A single tree is
easy to understand but unstable — small data changes can reshape it
completely.

A **random forest** grows many trees (here `ntree = 500`), each on a
random bootstrap sample of the data and a random subset of predictors,
then averages their predictions. Averaging reduces variance and gives a
more robust model. Two things fall out of the `print()` output:

- **Mean of squared residuals (MSE)** — the average squared prediction
  error.
- **% Var explained** — the proportion of variance the model explains,
  computed *out-of-bag* (on the data each tree did not see during
  training). This is an honest, built-in estimate of generalisation.
:::

**Key parameters:** - `ntree = 500`: Number of trees.
More trees reduce variance but increase computation time.
- `seed = 222`: Ensures reproducible random splits.

**Expected output:** A regression `randomForest` object summary showing % variance explained and mean squared error (mean of squared residuals).

::: callout-note
## Reading the result

Because our six risk-factor covariates were simulated independently of
`incidence_rate` (@sec-simulate), there is no underlying mathematical
pattern to find. The out-of-bag **% variance explained is only about
1.3%** — essentially zero. A model that simply predicted the average
rate for every country would do about as well. This is the expected,
honest outcome for a signal-free dataset: the algorithm ran correctly
and correctly found nothing.
:::

```{r rf-var-imp}
rf_var_imp <- caret::varImp(rf_model) |> 
  as_tibble(rownames = "Variables")

# Build the plot
ggplot(rf_var_imp, aes(x = reorder(Variables, abs(Overall)), y = Overall)) +
  geom_col(aes(fill = Overall >= 0), width = 0.7, alpha = 0.75) +
  scale_fill_manual(values = c("TRUE" = "steelblue", "FALSE" = "firebrick4"),
                    labels = c("Negative", "Positive"),
                    name = "Direction") +
  geom_line(aes(group = 1), colour = "darkslategrey", linewidth = 1) +
  geom_point(colour = "darkslategrey", size = 3) +
  geom_point(colour = "white", size = 1.5) +
  geom_text(aes(label = round(Overall,2)),
            hjust = if_else(rf_var_imp$Overall >= 0, -0.15, 1.15),
            vjust = 0.5,
            fontface = "bold",
            size = 3) +
  coord_flip() +
  labs(
    title = "Variable Importance Plot",
    y = "Importance Score", x = ""
  ) +
  theme_minimal(base_size = 12) +
  expand_limits(y = c(min(rf_var_imp$Overall) - 1, max(rf_var_imp$Overall) + 1))
```

::: callout-note
## Reading the importance plot

`caret::varImp(rf_model)` extracts the importance of each predictor —
for random forests, the percentage increase in MSE when that variable is
permuted (shuffled). A **larger positive** value means the model relies
on it more; a **negative** value means shuffling that variable *improved*
the model, which happens when a variable contributes nothing but noise.

Here `TOTPOP` is mildly elevated, but the mixed positive/negative bars
show the model is fitting random artefacts rather than a biological
signal.
:::

The variable importance plot shows a mildly elevated `TOTPOP` variable as the most important, given its earlier partial correlation in @sec-diagnose.
This bar chart with negative importance clearly shows that the model is fitting random artefacts rather than true biological signals.

------------------------------------------------------------------------

## Train an XGBoost Model {#sec-xgb}

**Purpose:** Apply gradient boosting — a sequential ensemble that corrects errors of previous trees — for potentially higher accuracy.

```{r train-xgb}
# XGB requires a numeric matrix but train_xgb() handles that internally
set.seed(222)
xgb_model <- train_xgb(
  data = ml_data,
  formula = rf_formula,  # same predictors
  nrounds = 100,
  max_depth = 4,
  learning_rate = 0.1
)

print(xgb_model)
```

::: callout-note
## What XGBoost is

Where a random forest grows many trees *independently* and averages
them, **gradient boosting** grows trees *sequentially*: each new tree is
fitted to the errors left by the previous ones. The result is a strong
model built from many weak ones.

The key hyperparameters are:

- `nrounds` — the number of boosting rounds (trees added).
- `max_depth` — how deep each tree may grow; smaller = less overfitting.
- `learning_rate` (η) — how much each tree's contribution is shrunk;
  smaller means slower but often better learning (typical 0.01–0.3).

`train_xgb()` handles the bookkeeping of turning the formula into a
numeric matrix for you.
:::

**Key parameters:** - `nrounds`: Number of boosting iterations.
Early stopping is recommended for large datasets.
- `max_depth`: Controls tree complexity; lower values reduce overfitting.
- `learning_rate` (η): Shrinks contribution of each tree; typical range 0.01–0.3.

**Expected output:** An `xgb.Booster` object showing the model summary confirming the number of boosting rounds and features used.
Unlike the random forest summary, xgboost does not print the residual/variance diagnostic directly; we compute them explicitly in @sec-eval.

```{r xgb-var-imp}
xgb_var_imp <- xgb.importance(xgb_model) |> 
  as_tibble()

# Build the plot
ggplot(xgb_var_imp, aes(x = reorder(Feature, Gain), y = Gain)) +
  geom_col(width = 0.7, alpha = 0.75, fill = "steelblue") +
  geom_line(aes(group = 1), colour = "darkslategrey", linewidth = 1) +
  geom_point(colour = "darkslategrey", size = 3) +
  geom_point(colour = "white", size = 1.5) +
  geom_text(aes(label = round(Gain, 2)),
            hjust = -0.5,
            vjust = 0.5,
            fontface = "bold",
            size = 3) +
  scale_y_continuous(expand = expansion(mult = c(0.05, 0.1))) +
  coord_flip() +
  labs(
    title = "Variable Importance (Gain)",
    y = "Gain Score", x = ""
  ) +
  theme_minimal(base_size = 12)
```

::: callout-note
## Reading `xgb.importance()`

`xgb.importance()` ranks predictors by three measures:

- **Gain** — the average improvement in the loss function each variable
  contributes when used in a split. This is the headline importance.
- **Cover** — how many observations pass through splits on that variable.
- **Frequency** — how often the variable is used in a split.

Unlike random-forest importance, Gain is always non-negative, which is
why this plot has no bars below zero.
:::

The variable importance bar chart is ranked by `Gain`, the average improvement in the loss function each variable contributes when used in a split.
As with the random forest, `TOTPOP` has the highest gain but it is almost similar to some of the covariates.
Also there is no negative importance in this plot.

------------------------------------------------------------------------

## Train a Support Vector Regression Model {#sec-svr}

**Purpose:** Fit a kernel-based model that is robust to high-dimensional data and captures non-linear boundaries via the radial basis function (RBF), after scaling the predictors.

```{r train-svr}
# Scale predictors before training
predictors <- c("smoking_prev", "obesity_prev", "diabetes_prev",
                "alcohol_per_capita", "family_history_prev",
                "urbanization_pct", "TOTPOP")
ml_data_scaled <- ml_data |> 
  mutate(across(all_of(predictors), ~ as.numeric(scale(.x))))


svr_model <- train_svr(
  data = ml_data,
  formula = rf_formula
)

print(svr_model)
```

::: callout-note
## What Support Vector Regression is

SVR fits a function that is *flat* — within a margin (ε) it tolerates
errors, and it only cares about points outside that margin (the
**support vectors**). The **radial (RBF) kernel** lets it capture
non-linear relationships by mapping points into a higher-dimensional
space without doing the maths explicitly.

Two things to note from the summary:

- **Number of support vectors** — here about 49 of the 54 rows, which is
  typical for small, noisy datasets: the model cannot find a simple
  structure, so nearly every point ends up supporting the fit.
- The predictors are scaled first (`scale()` gives each a mean of 0 and
  SD of 1), because the RBF kernel is distance-based.
:::

**Key parameters:** - `train_svr()` uses the radial kernel by default (`kernel = "radial"` in `e1071::svm()`).
- Tuning `cost` and `gamma` via cross-validation (e.g., `caret::train()`) is recommended for production models.
- `scale()` standardises each predictor to a mean of 0, SD of 1.
- `as.numeric` strips the matrix class of the `scale()` output into a clean numeric column.

**Expected output:** An `svm` summary showing the number of support vectors and model parameters.
The number of support vectors is fairly high for a 54-row dataset, which is normal for small datasets.

------------------------------------------------------------------------

## Evaluate Model Performance {#sec-eval}

**Purpose:** Compare models using standardized accuracy metrics: RMSE (Root Mean Squared Error), MAE (Mean Absolute Error), and R² (coefficient of determination).

```{r eval-models}
# Generate in-sample predictions from each model and compute RMSE/MAE/R2
# using caret::postResample()

# Random Forest predictions
rf_preds <- predict(rf_model, newdata = ml_data)
rf_metrics <- postResample(pred = rf_preds, obs = ml_data$incidence_rate)

# XGBoost predictions
xgb_preds <- predict(xgb_model, newdata = as.matrix(ml_data[, predictors]))
xgb_metrics <- postResample(pred = xgb_preds, obs = ml_data$incidence_rate)

# SVR predictions (on the scaled predictor set used for training)
svr_preds <- predict(svr_model, newdata = ml_data_scaled)
svr_metrics <- postResample(pred = svr_preds, obs = ml_data$incidence_rate)

# Combine into a comparison table
model_eval <- data.frame(
  Model = c("Random Forest", "XGBoost", "SVR"),
  RMSE = c(rf_metrics["RMSE"], xgb_metrics["RMSE"], svr_metrics["RMSE"]),
  MAE  = c(rf_metrics["MAE"], xgb_metrics["MAE"], svr_metrics["MAE"]),
  Rsquared = c(rf_metrics["Rsquared"], xgb_metrics["Rsquared"], svr_metrics["Rsquared"])
)

model_eval[order(model_eval$RMSE), ]
```

::: callout-note
## The three metrics

- **RMSE** — square root of the mean squared error; penalises large
  errors more, and is in the same units as the outcome.
- **MAE** — mean absolute error; the average size of an error, easy to
  interpret.
- **R²** — the proportion of variance explained (1 = perfect, 0 = no
  better than predicting the mean).

`caret::postResample(pred, obs)` computes all three from predictions and
the true values.
:::

**Key parameters:** - `predict()`: Requires a `newdata` to supply the predictor variables to generate predictions for a particular model.
- `postResample()` compares the predicted values against ground truth (the target variable, `incidence_rate`) to generate standard regression performance metrics.

**Expected output:** A three-row data frame ranking the models.

::: callout-warning
## In-sample ≠ generalisation

These are **in-sample (training-set)** metrics: every model is evaluated
on the very data it was trained on. A high R² here means the model
*memorised* the training data, not that it learned a real relationship.
Tree-based models in particular (especially XGBoost with enough rounds)
can fit training data almost perfectly even when the true relationship is
weak or non-existent — exactly the scenario we have with our uncorrelated
simulated covariates. **A high in-sample R² is not evidence of a real
relationship**; it is a reminder of why we need held-out cross-validation,
which the next section does.
:::

------------------------------------------------------------------------

## Honest Model Evaluation with `caret` {#sec-honest-eval}

**Purpose:** Replace the in-sample metrics with `caret::train`, which provides built-in cross-validation, consistent evaluation across all three algorithms, and avoids the in-sample inconsistencies discovered above.

```{r caret-cv}
# Refit Random Forest, XGBoost and Support Vector Regression
# using train() from caret via the same repeated 5-fold cross-validation
set.seed(222)

cv_control <- trainControl(method = "repeatedcv", number = 5, 
                           repeats = 5, savePredictions = "final")

rf_cv <- train(
  rf_formula, data = ml_data, method = "rf", trControl = cv_control
)
print(rf_cv)

cat("\n")
svr_cv <- train(
  rf_formula, data = ml_data, method = "svmRadial",
  trControl = cv_control, 
  preProcess = c("center", "scale")  # caret scales predictors per fold
)
print(svr_cv)

# Manually cross-validate XGBoost using xgb.cv
# (current version of xgboost breaks with caret::train)
xgb_matrix_full <- xgb.DMatrix(
  data  = as.matrix(ml_data[, predictors]),
  label = ml_data$incidence_rate
)

xgb_cv <- xgb.cv(
  data = xgb_matrix_full,
  params = list(
    objective = "reg:squarederror",
    max_depth = 4,
    eta = 0.1
  ),
  metrics = list("rmse","mae"),
  nrounds = 100,
  nfold = 5,
  showsd = TRUE,
  prediction = TRUE,
  early_stopping_rounds = 10, # stops at best iter
  verbose = 0
)

# Best cross-validated metrics across boosting rounds
cat("\n======= XGBoost Metrics =========")
cat("\nRMSE:", min(xgb_cv$evaluation_log$test_rmse_mean))
cat("\nMAE:", min(xgb_cv$evaluation_log$test_mae_mean))
cat("\nRsquared:", cor(ml_data$incidence_rate, as.vector(xgb_cv$cv_predict$pred))^2)
```

::: callout-note
## What cross-validation does

**k-fold cross-validation** splits the data into *k* parts, repeatedly
trains on *k−1* parts, and evaluates on the held-out part. Every row is
predicted by a model that never saw it, so the resulting metric is an
honest estimate of performance on *new* data — the opposite of the
in-sample metrics above.

- `trainControl(method = "repeatedcv", number = 5, repeats = 5)` does
  5-fold CV repeated 5 times with different random splits (25 fits per
  candidate model), reducing the noise in the estimate.
- `method = "rf"` and `"svmRadial"` are `caret`'s model codes for random
  forest and radial-kernel SVM. `preProcess = c("center", "scale")`
  scales predictors *within each fold* — done inside CV so scaling is not
  contaminated by the held-out data.
- The latest `xgboost` has a compatibility issue with `caret::train`, so
  XGBoost is cross-validated directly with `xgb.cv()`. `early_stopping_rounds`
  halts training when the held-out error stops improving.
:::

**Key parameters:** - `trainControl()` with `"repeatedcv"` performs 5-fold cross-validation repeated 5 times with different random fold assignments each time.
- `method = "rf"`, `"svmRadial"` are `caret`'s standardised model-methods for random forest and SVM.
- `preProcess` standardises the predictors within each fold for SVM.

**Expected output:** Cross-validated out-of-fold metrics for random forest and SVR (via `caret::train`) and for XGBoost (manually calculated).

::: callout-note
## The honest result

With no signal in the covariates, the cross-validated R² collapses toward
zero for all three models (roughly 0.15 for RF, 0.07 for SVR, and ≈ 0 for
XGBoost) — a stark contrast to the near-perfect in-sample fits. This
collapse *is* the lesson: never trust a training-set score alone.
:::

------------------------------------------------------------------------

## Identify Spatial Patterns and Risk Factors {#sec-patterns}

**Purpose:** Use the trained models' predictions to check whether any spatial pattern emerges, to inform health decision-making.

```{r ml-patterns}
# Attach out-of-fold predictions back onto the spatial object
rf_oof <- rf_cv$pred |> 
  group_by(rowIndex) |> 
  summarise(pred_rf = mean(pred), .groups = "drop") |> 
  arrange(rowIndex)

svr_oof <- svr_cv$pred |> 
  group_by(rowIndex) |> 
  summarise(pred_svr = mean(pred), .groups = "drop") |> 
  arrange(rowIndex)

xgb_oof <- tibble(
  rowIndex = seq_len(nrow(ml_data)),
  pred_xgb = as.vector(xgb_cv$cv_predict$pred)
)

africa_health <- africa_health |> 
  mutate(
    pred_rf  = rf_oof$pred_rf,
    pred_xgb = xgb_oof$pred_xgb,
    pred_svr = svr_oof$pred_svr
  )

# Observed vs Predicted diagnostic plot via mlspatial
plot_obs_vs_pred(
  observed = africa_health$incidence_rate,
  predicted = africa_health$pred_rf,
  title = "Random Forest CV: Observed vs Predicted"
)

plot_obs_vs_pred(
  observed = africa_health$incidence_rate,
  predicted = africa_health$pred_svr,
  title = "SVR CV: Observed vs Predicted"
)

plot_obs_vs_pred(
  observed = africa_health$incidence_rate,
  predicted = africa_health$pred_xgb,
  title = "XGB CV: Observed vs Predicted"
)

# Map predictions vs observed
obs_map <- tm_shape(africa_health) +
  tm_fill(
    fill = "incidence_rate",
    fill.scale = tm_scale(values = "brewer.yl_or_rd"),
    fill.legend = tm_legend(title = "Observed")
  ) +
  tm_borders() 

pred_rf_map <- tm_shape(africa_health) +
  tm_fill(
    fill = "pred_rf",
    fill.scale = tm_scale(values = "brewer.yl_or_rd"),
    fill.legend = tm_legend(title = "Predicted (RF)")
  ) +
  tm_borders() 

pred_xgb_map <- tm_shape(africa_health) +
  tm_fill(
    fill = "pred_xgb",
    fill.scale = tm_scale(values = "brewer.yl_or_rd"),
    fill.legend = tm_legend(title = "Predicted (XGB)")
  ) +
  tm_borders() 

pred_svr_map <- tm_shape(africa_health) +
  tm_fill(
    fill = "pred_svr",
    fill.scale = tm_scale(values = "brewer.yl_or_rd"),
    fill.legend = tm_legend(title = "Predicted (SVR)")
  ) +
  tm_borders() 

tmap_arrange(obs_map, pred_rf_map, pred_xgb_map, pred_svr_map, ncol = 2)
```

::: callout-note
## Reading this section

- **Out-of-fold predictions** are the predictions made on each row while
  it was held out during CV. `rf_cv$pred` stores them; we
  `group_by(rowIndex)` and average over the repeated folds to get one
  prediction per country, then re-attach them to `africa_health` by row
  order (the same trick used to drop the geometry earlier).
- `plot_obs_vs_pred()` draws observed vs predicted with a dashed 1:1
  line: points hugging that line indicate a good model. For our
  signal-free data the points scatter widely around it.
- The four maps compare the observed map with each model's predicted
  map. If a model had learned a real spatial pattern, its predictions
  would reproduce the observed map's structure.
:::

**Expected output:** - A scatterplot with a 1:1 line and Pearson's r.
- Side-by-side maps showing how well the RF, XGB and SVR models reproduce spatial patterns.

Across every method in this module — Random Forest, XGBoost, SVR — the consistent finding is that our six simulated lifestyle/genetic covariates carry no real signal.
This is the intended result of a **teaching dataset**: it demonstrates that the entire ML pipeline (data prep → training → tuning → evaluation → spatial pattern mapping) runs correctly and produces internally consistent, honestly interpreted null results.
When you apply this exact pipeline to real covariate data in your own research, this is the standard of scrutiny that should accompany any reported results.

------------------------------------------------------------------------

## Practical Exercise {#sec-m5-exercise}

Consolidate the module with a few tasks, then check your answers against the solution.

1. Train a Random Forest with `ntree = 200` and a different seed.
   Report the % variance explained.
2. Use `xgb.importance()` to list the **top three** predictors by Gain.
3. Compute the in-sample R² of the SVR model with `postResample()`.
4. *(Challenge)* Re-run the `caret` random-forest cross-validation with
   `number = 10` (10-fold) and compare the best RMSE with the 5-fold
   result above.

::: {.callout-note collapse="true"}
## Model solution

```r
library(sf)
library(dplyr)
library(tidyr)
library(mlspatial)
library(randomForest)
library(xgboost)
library(e1071)
library(caret)

data(africa_shp, package = "mlspatial")
data(panc_incidence, package = "mlspatial")
africa_health <- join_data(
  africa_shp |> mutate(NAME = gsub("Côte d'Ivoire", "Cote d'Ivoire", NAME)),
  panc_incidence, by = "NAME"
)
africa_health <- st_make_valid(africa_health)

set.seed(898)
n <- nrow(africa_health)
africa_health <- africa_health |>
  mutate(
    incidence_rate = (incidence / TOTPOP) * 100000,
    smoking_prev = round(runif(n, 8, 28), 1),
    obesity_prev = round(runif(n, 5, 30), 1),
    diabetes_prev = round(runif(n, 2, 11), 1),
    alcohol_per_capita = round(runif(n, 0.5, 12), 1),
    family_history_prev = round(runif(n, 3, 10), 1),
    urbanization_pct = round(runif(n, 15, 85), 1)
  )

ml_data <- africa_health |> st_drop_geometry() |>
  select(incidence_rate, smoking_prev, obesity_prev, diabetes_prev,
         alcohol_per_capita, family_history_prev, urbanization_pct, TOTPOP) |>
  drop_na()

rf_formula <- incidence_rate ~ smoking_prev + obesity_prev + diabetes_prev +
  alcohol_per_capita + family_history_prev + urbanization_pct + TOTPOP
predictors <- c("smoking_prev", "obesity_prev", "diabetes_prev",
                "alcohol_per_capita", "family_history_prev",
                "urbanization_pct", "TOTPOP")

# 1. Random forest with ntree = 200, new seed
rf2 <- train_rf(data = ml_data, formula = rf_formula, ntree = 200, seed = 4815)
tail(rf2$rsq, 1) * 100      # % variance explained (near 0)

# 2. Top-3 XGBoost predictors by Gain
set.seed(222)
xgb_model <- train_xgb(data = ml_data, formula = rf_formula,
                       nrounds = 100, max_depth = 4, learning_rate = 0.1)
xgb.importance(xgb_model) |> head(3)

# 3. In-sample SVR R-squared
ml_data_scaled <- ml_data |>
  mutate(across(all_of(predictors), ~ as.numeric(scale(.x))))
svr_model <- train_svr(data = ml_data, formula = rf_formula)
postResample(predict(svr_model, newdata = ml_data_scaled),
             ml_data$incidence_rate)["Rsquared"]

# 4. 10-fold RF cross-validation
set.seed(222)
cv10 <- trainControl(method = "repeatedcv", number = 10, repeats = 5,
                     savePredictions = "final")
rf_cv10 <- train(rf_formula, data = ml_data, method = "rf", trControl = cv10)
min(rf_cv10$results$RMSE)
```
:::

------------------------------------------------------------------------

## Module 5 Checklist {#se-m5-check .unnumbered}

- [ ] Prepared clean ML data with `st_drop_geometry()`
- [ ] Trained Random Forest, XGBoost, and SVR models
- [ ] Evaluated all three models with `postResample()`
- [ ] Compared RMSE, MAE, and R² across algorithms
- [ ] Re-evaluated honestly with cross-validation
- [ ] Mapped predictions and identified top risk factors
- [ ] Practical exercise completed

**Next:** Proceed to [Module 6: Integrated Application from Data to Decision](06-integration.qmd).
