---
editor: 
  markdown: 
    wrap: sentence
---

# Module 5: Machine Learning for Spatial Epidemiology (Showcasing `mlspatial`) {#sec-module5}

This module showcases the machine learning capabilities of `mlspatial`.
You will prepare data, train three algorithms (Random Forest, XGBoost, Support Vector Regression), evaluate their performance, and map predictions to identify spatial risk patterns.

```{r setup-m5}
#| include: false
library(sf)
library(terra)
library(dplyr)
library(tidyr)
library(mlspatial)
library(tmap)
library(randomForest)
library(xgboost)
library(e1071)
library(caret)

data(africa_shp, package = "mlspatial")
data(panc_incidence, package = "mlspatial")
africa_health <- join_data(
  africa_shp |> mutate(
    across(c(NAME, COUNTRYAFF),
           ~ replace_values(.x, "Côte d'Ivoire" ~ "Cote d'Ivoire"))),
  panc_incidence, by = "NAME"
)
africa_health <- st_make_valid(africa_health)

# Add simulated environmental data
set.seed(898)  
n <- nrow(africa_health)
africa_health <- africa_health |> 
  mutate(
    incidence_rate = (incidence/TOTPOP)*100000,
    smoking_prev = round(runif(n, min = 8, max = 28), 1),
    obesity_prev = round(runif(n, min = 5, max = 30), 1),
    diabetes_prev = round(runif(n, min = 2,  max = 11), 1),
    alcohol_per_capita = round(runif(n, min = 0.5, max = 12), 1),
    family_history_prev = round(runif(n, min = 3,  max = 10), 1),
    urbanization_pct = round(runif(n, min = 15, max = 85), 1)
  )
```

------------------------------------------------------------------------

## Introduction to `mlspatial` for ML {#sec-ml-intro}

**Purpose:** Understand what `mlspatial` provides and how it fits into a spatial epidemiology ML workflow, before writing any modelling code.

Traditional spatial epidemiology often requires switching between GIS software, R scripts, and Python notebooks.
`mlspatial` (Azeez & Noel, 2025) packages together spatial data handling (`sf`), thematic mapping(`tmap`), spatial autocorrelation diagnostics (`spdep`), and machine learning (`randomforest`, `xgboost` and `e1071` ) into a single applied workflow aimed at spatial epidemiologists.

The package provides a number of convenient functions and leans on well-established packages for this workflow.
In this module we will train three ML algorithms—Random Forest, XGBoost, and Support Vector Regression—to predict pancreatic cancer incidence from our simulated covariates.

------------------------------------------------------------------------

## Prepare Data for ML Modelling {#sec-ml-prep}

**Purpose:** Prepare our `sf` object into a clean, geometry-free data frame for machine learning algorithms.
We drop the geometry column, select only relevant columns and check for missing values and decide whether scaling is needed for each algorithm.

```{r ml-prep}
# Drop geometry and select complete cases
ml_data <- africa_health |> 
  st_drop_geometry() |> 
  select(
    incidence_rate,                    # outcome
    smoking_prev, obesity_prev, diabetes_prev,
    alcohol_per_capita, family_history_prev,
    urbanization_pct, TOTPOP                 
  ) |> 
  drop_na()

# Quick structural check
str(ml_data)
sum(is.na(ml_data))  
```

**Key parameters:** - `st_drop_geometry()`: ML algorithms cannot use sf geometry columns; we keep the spatial link via row order for later mapping.
- `drop_na()`: XGBoost and SVR do not tolerate missing values.

**Expected output:** A plain `data.frame` (no `sf`/`data.fram` dual class) with remaining rows after removing `NAs` and 8 columns.

**On Scaling:** Unlike linear/distance based models, tree-based methods (**Random Forest, XGBoost**) are scale-invariant so standardising predictors has no effect on their fit.
**SVR** however is distance sensitive (its algorithm computes similarity based on Euclidean distance in feature space), so we scale predictors only immediately before the SVR model fit in @sec-svr, rather than scaling the whole `ml_data` object upfront.

------------------------------------------------------------------------

## Train a Random Forest Model {#sec-rf}

**Purpose:** Use an ensemble of decision trees to capture non-linear relationships between predictors and `incidence_rate` and inspect variable importance

```{r train-rf}
# Train Random Forest regression model
rf_formula <- incidence_rate ~ smoking_prev + obesity_prev + diabetes_prev +
  alcohol_per_capita + family_history_prev + urbanization_pct + TOTPOP

rf_model <- mlspatial::train_rf(
  data = ml_data,
  formula = rf_formula,
  ntree = 500,
  seed = 222
)

print(rf_model)
```

**Key parameters:** - `ntree = 500`: Number of trees.
More trees reduce variance but increase computation time.
- `seed = 222`: Ensures reproducible random splits.

**Expected output:** A regression `randomForest` object summary showing % variance explained and mean squared error (mean of squared residuals).

Because our six-risk factor covariates were simulated independently of `incidence_rate` (@sec-simulate) , there is absolutely no underlying mathematical pattern to find.
The negative explained variance means this random forest model performs worse than a model that simply guesses the average incidence rate for every country.

```{r rf-var-imp}
rf_var_imp <- caret::varImp(rf_model) |> 
  as_tibble(rownames = "Variables")

# Build the plot
ggplot(rf_var_imp, aes(x = reorder(Variables, abs(Overall)), y = Overall)) +
  geom_col(aes(fill = Overall >= 0), width = 0.7, alpha = 0.75) +
  scale_fill_manual(values = c("TRUE" = "steelblue", "FALSE" = "firebrick4"),
                    labels = c("Negative", "Positive"),
                    name = "Direction") +
  geom_line(aes(group = 1), colour = "darkslategrey", linewidth = 1) +
  geom_point(colour = "darkslategrey", size = 3) +
  geom_point(colour = "white", size = 1.5) +
  geom_text(aes(label = round(Overall,2)),
            hjust = if_else(rf_var_imp$Overall >= 0, -0.15, 1.15),
            vjust = 0.5,
            fontface = "bold",
            size = 3) +
  coord_flip() +
  labs(
    title = "Variable Importance Plot",
    y = "Importance Score", x = ""
  ) +
  theme_minimal(base_size = 12) +
  expand_limits(y = c(min(rf_var_imp$Overall) - 1, max(rf_var_imp$Overall) + 1))
```

The variable importance plot shows a mildly elevated `TOTPOP` variable as the most important given its earlier partial correlation in @sec-diagnose.
This bar chart with -negative importance clearly shows that the model is fitting random artefacts than true biological signals.

------------------------------------------------------------------------

## Train an XGBoost Model {#sec-xgb}

**Purpose:** Apply gradient boosting—a sequential ensemble that corrects errors of previous trees—for potentially higher accuracy.

```{r train-xgb}
# XGB requires a numeric matrix but train_xgb() handles that internally
set.seed(222)
xgb_model <- train_xgb(
  data = ml_data,
  formula = rf_formula,  # same predictors
  nrounds = 100,
  max_depth = 4,
  learning_rate = 0.1
)

print(xgb_model)
```

**Key parameters:** - `nrounds`: Number of boosting iterations.
Early stopping is recommended for large datasets.
- `max_depth`: Controls tree complexity; lower values reduce overfitting.
- `learning_rate` (η): Shrinks contribution of each tree; typical range 0.01–0.3.

**Expected output:** An `xgb.Booster` object showing the model summary confirming the number of boosting rounds and features used.
Unlike the random forest summary, xgboost does not print the residual/variance diagnostic directly.
we compute them explicitly in @sec-eval

```{r xgb-var-imp}
xgb_var_imp <- xgb.importance(xgb_model) |> 
  as_tibble()

# Build the plot
ggplot(xgb_var_imp, aes(x = reorder(Feature, Gain), y = Gain)) +
  geom_col( width = 0.7, alpha = 0.75, fill = "steelblue") +
  geom_line(aes(group = 1), colour = "darkslategrey", linewidth = 1) +
  geom_point(colour = "darkslategrey", size = 3) +
  geom_point(colour = "white", size = 1.5) +
  geom_text(aes(label = round(Gain, 2)),
            hjust = -0.5,
            vjust = 0.5,
            fontface = "bold",
            size = 3) +
  scale_y_continuous(expand = expansion(mult = c(0.05, 0.1))) +
  coord_flip() +
  labs(
    title = "Variable Importance (Gain)",
    y = "Gain Score", x = ""
  ) +
  theme_minimal(base_size = 12)
```

The variable importance bar chart is ranked by `Gain` , the average improvement in the loss function each variable contributes when used in a split.
As with the random forest `TOTPOP` has the highest `gain` but it is almost similar to some of the covariates.
Also there is no negative importance in this plot.

------------------------------------------------------------------------

## Train a Support Vector Regression Model {#sec-svr}

**Purpose:** Fit a kernel-based model that is robust to high-dimensional data and captures non-linear boundaries via the radial basis function (RBF) after scaling the predictors.

```{r train-svr}
# Scale predictors before training
predictors <- c("smoking_prev", "obesity_prev", "diabetes_prev",
                "alcohol_per_capita", "family_history_prev",
                "urbanization_pct", "TOTPOP")
ml_data_scaled <- ml_data |> 
  mutate(across(all_of(predictors), ~ as.numeric(scale(.x))))


svr_model <- train_svr(
  data = ml_data,
  formula = rf_formula
)

print(svr_model)
```

**Key parameters:** - `train_svr()` uses the radial kernel by default (`kernel = "radial"` in `e1071::svm()`).
- Tuning `cost` and `gamma` via cross-validation (e.g., `caret::train()`) is recommended for production models.
- `scale()` standardises each predictor to a mean of 0, SD of 1.
- `as.numeric` strips the matrix class of the `scale` output into a clean numeric column in the dataframe.

**Expected output:** An `svm` summary showing number of support vectors and model parameters.
The number of support vectors is fairly high for a 54 row data, which is normal for small datasets.

------------------------------------------------------------------------

## Evaluate Model Performance {#sec-eval}

**Purpose:** Compare models using standardized accuracy metrics: RMSE (Root Mean Squared Error), MAE (Mean Absolute Error), and R² (coefficient of determination).

```{r eval-models}
# Generate in sample predictions from each modeland compute RMSE/MAE/R2
# using caret::postResample()

# Random Forest predictions
rf_preds <- predict(rf_model, newdata = ml_data)
rf_metrics <- postResample(pred = rf_preds, obs = ml_data$incidence_rate)

# XGBoost predictions
xgb_preds <- predict(xgb_model, newdata = as.matrix(ml_data[, predictors]))
xgb_metrics <- postResample(pred = xgb_preds, obs = ml_data$incidence_rate)

# SVR predictions (on the scaled predictor set used for training)
svr_preds <- predict(svr_model, newdata = ml_data_scaled)
svr_metrics <- postResample(pred = svr_preds, obs = ml_data$incidence_rate)

# Combine into a comparison table
model_eval <- data.frame(
  Model = c("Random Forest", "XGBoost", "SVR"),
  RMSE = c(rf_metrics["RMSE"], xgb_metrics["RMSE"], svr_metrics["RMSE"]),
  MAE  = c(rf_metrics["MAE"], xgb_metrics["MAE"], svr_metrics["MAE"]),
  Rsquared = c(rf_metrics["Rsquared"], xgb_metrics["Rsquared"], svr_metrics["Rsquared"])
)

model_eval[order(model_eval$RMSE), ]
```

**Key parameters:** - `predict()`: Requires a `newdata` to supply the predictor variables to genrate predictions for a particular model.
- `postResample` compares the predicted values against ground truth (target variable - `incidence_rate`) to generate standard regression performance metrics.

**Expected output:** A three-row data frame ranking models.

**Important Point to Note**: These are in-sample (training-set) metrics, every model is evaluated on the same data it was trained on, so R2 here reflects how well each model *fit* the training data, not how well it would *generalise* to new countries.
Tree-based models in particular (especially XGBoost with enough rounds) can fit training data almost perfectly (memorise) even when the true relationship is weak or nonexixstent - exactly the scenario we have here with our uncorrelated simulated covariates.
**A high in sample R2 here should not be read as evidence the model has learned a real relationship**, it is a reminder of why held-out cross validation

---

## Honest Model Evaluation with `caret`

**Purpose:** replace `mlspatial`'s `train_*`/`eval_model()` workflow with `caret::train` which provides built-in cross-validation, consistent evaluation across all three algorithms and avoids the in-sample inconsistencies discovered above.

```{r}
# Refit Random Forest, XGBoost and Support Vector Regression
# using train() from caret via the same repeated 5-fold cross-validation
set.seed(222)

cv_control <- trainControl(method = "repeatedcv", number = 5, 
                           repeats = 5, savePredictions = "final")

rf_cv <- train(
  rf_formula, data = ml_data, method = "rf", trControl = cv_control
)
print(rf_cv)

svr_cv <- train(
  rf_formula, data = ml_data, method = "svmRadial",
  trControl = cv_control, 
  preProcess = c("center", "scale")  # caret scales predictors per fold
)
print(svr_cv)

# Manually cross-validate XGBoost using xgb.cv
# current version of xgboost breaks with caret::train 
xgb_matrix_full <- xgb.DMatrix(
  data  = as.matrix(ml_data[, predictors]),
  label = ml_data$incidence_rate
)

xgb_cv <- xgb.cv(
  data = xgb_matrix_full,
  params = list(
    objective = "reg:squarederror",
    max_depth = 4,
    eta = 0.1
  ),
  metrics = list("rmse","mae"),
  nrounds = 100,
  nfold = 5,
  showsd = TRUE,
  prediction = TRUE,
  early_stopping_rounds = 10, #stops at best iter
  verbose = 0
)

# Best cross-validated metrics across boosting rounds
cat("\n======= XGBoost Metrics =========")
cat("\nRMSE:", min(xgb_cv$evaluation_log$test_rmse_mean))
cat("\nMAE:", min(xgb_cv$evaluation_log$test_mae_mean))
cat("\nRsquared:", cor(ml_data$incidence_rate, as.vector(xgb_cv$cv_predict$pred))^2)
```

**Key parameter:** -`trainControl()` with "repeatedcv" performs a 5-fold cross validation repeated 5 times with different random fold assignments each time.
-`method="rf"`, `"svmRadial"` is `caret`'s standardised model-methods for random forest and svm.
There is a current compatibility issue with the latest `xgboost` package and `caret` so we use `xgb.cv` to perform the cross validation for the xgboost model directly via the `xgboost` package.
-`Preprocess` standardises the predictors within each fold for svm.

**Expected output:** the cross validate out of fold metrics fro random forest and support vector regression via `caret::train` and that for xgboost manually calculated.
Given everything established in this module about our simulated, signal free covariates, all the three model's R2 values now reflect an honest estimation of model performance in stark contrast to the earlier in-sample estimates.

------------------------------------------------------------------------

## Identify Spatial Patterns and Risk Factors {#sec-patterns}

**Purpose:** Use trained model's predictions to check whether an spatial pattern emerges in prediction to inform health decision making.

```{r ml-patterns}
# Attach out of fold predictions back onto the spatial object
rf_oof <- rf_cv$pred |> 
  group_by(rowIndex) |> 
  summarise(pred_rf = mean(pred), .groups = "drop") |> 
  arrange(rowIndex)

svr_oof <- svr_cv$pred |> 
  group_by(rowIndex) |> 
  summarise(pred_svr = mean(pred), .groups = "drop") |> 
  arrange(rowIndex)

xgb_oof <- tibble(
  rowIndex = seq_len(nrow(ml_data)),
  pred_xgb = as.vector(xgb_cv$cv_predict$pred)
)

africa_health <- africa_health |> 
  mutate(
    pred_rf  = rf_oof$pred_rf,
    pred_xgb = xgb_oof$pred_xgb,
    pred_svr = svr_oof$pred_svr
  )

# Observed vs Predicted diagnostic plot via mlspatial
plot_obs_vs_pred(
  observed = africa_health$incidence_rate,
  predicted = africa_health$pred_rf,
  title = "Random Forest CV: Observed vs Predicted"
)

plot_obs_vs_pred(
  observed = africa_health$incidence_rate,
  predicted = africa_health$pred_svr,
  title = "SVR CV: Observed vs Predicted"
)


# Map predictions vs observed
obs_map <- tm_shape(africa_health) +
  tm_fill(
    fill = "incidence_rate",
    fill.scale = tm_scale(values = "brewer.yl_or_rd"),
    fill.legend = tm_legend(title = "Observed")
  ) +
  tm_borders() 

pred_rf_map <- tm_shape(africa_health) +
  tm_fill(
    fill = "pred_rf",
    fill.scale = tm_scale(values = "brewer.yl_or_rd"),
    fill.legend = tm_legend(title = "Predicted (RF)")
  ) +
  tm_borders() 

pred_xgb_map <- tm_shape(africa_health) +
  tm_fill(
    fill = "pred_xgb",
    fill.scale = tm_scale(values = "brewer.yl_or_rd"),
    fill.legend = tm_legend(title = "Predicted (XGB)")
  ) +
  tm_borders() 

pred_svr_map <- tm_shape(africa_health) +
  tm_fill(
    fill = "pred_svr",
    fill.scale = tm_scale(values = "brewer.yl_or_rd"),
    fill.legend = tm_legend(title = "Predicted (SVR)")
  ) +
  tm_borders() 

tmap_arrange(obs_map, pred_rf_map, pred_xgb_map, pred_svr_map, ncol = 2)
```

**Expected output:** - A scatterplot with a 1:1 line and Pearson's r.
- Side-by-side maps showing how well the RF, XGB and SVR model reproduces spatial patterns.

Across every method in this module - Random Forest, XGBoost, SVR, the consistent finding is that our six simulated lifestyle/genetic covariates carry no real signal.
This is the intended result of a **teaching dataset**: it demonstrates that the entire ML pipeline (data prep → training → tuning → evaluation → spatial pattern mapping) runs correctly and produces internally consistent honestly interpreted null results.
When you apply this exact pipeline to real covariate data in your own research, this is the standard of scrutiny that should accompany any reported results.

------------------------------------------------------------------------

## Module 5 Checklist {#se-m5-check .unnumbered}

- [ ] Prepared clean ML data with `st_drop_geometry()`
- [ ] Trained Random Forest, XGBoost, and SVR models
- [ ] Evaluated all three models with `eval_model()`
- [ ] Compared RMSE, MAE, and R² across algorithms
- [ ] Mapped predictions and identified top risk factors

**Next:** Proceed to [Module 6: Integrated Application from Data to Decision](06-integration.qmd).
