---
editor: 
  markdown: 
    wrap: sentence
---

# Module 6: Integrated Application from Data to Decision {#sec-module6}

This final module brings everything together.
You will overlay disease burden with health facility locations, calculate accessibility distances, generate reproducible reports, and reflect on the ethical dimensions of spatial epidemiology.

**What you will learn**

- How to combine a disease-burden map with service locations to spot underserved areas.
- How to compute the distance from each region to its nearest facility.
- How to turn your analysis into a reproducible report.
- The ethical obligations that come with mapping people and disease.

This module is the payoff of the whole tutorial: the earlier modules taught you *how* to import, map, model, and predict; this one shows you how those pieces combine into a decision-ready deliverable.

```{r setup-m6}
#| include: false
library(sf)
library(terra)
library(dplyr)
library(mlspatial)
library(tmap)
library(ggplot2)

data(africa_shp, package = "mlspatial")
data(panc_incidence, package = "mlspatial")
africa_health <- join_data(
  africa_shp |> mutate(
    NAME       = gsub("Côte d'Ivoire", "Cote d'Ivoire", NAME),
    COUNTRYAFF = gsub("Côte d'Ivoire", "Cote d'Ivoire", COUNTRYAFF)
  ),
  panc_incidence, by = "NAME"
)
africa_health <- st_make_valid(africa_health)

# Add simulated environmental data
set.seed(898)  
n <- nrow(africa_health)
africa_health <- africa_health |> 
  mutate(
    incidence_rate = (incidence/TOTPOP)*100000,
    smoking_prev = round(runif(n, min = 8, max = 28), 1),
    obesity_prev = round(runif(n, min = 5, max = 30), 1),
    diabetes_prev = round(runif(n, min = 2,  max = 11), 1),
    alcohol_per_capita = round(runif(n, min = 0.5, max = 12), 1),
    family_history_prev = round(runif(n, min = 3,  max = 10), 1),
    urbanization_pct = round(runif(n, min = 15, max = 85), 1)
  )
```

------------------------------------------------------------------------

## Overlay Disease Burden with Service Availability {#sec-overlay-decision}

**Purpose:** Identify under-served areas by superimposing disease burden on health facility locations.

```{r overlay-decision}
#| message: false

# Classify burden into tertiles (3 groups) for clearer visual communication
africa_health <- africa_health |>
  mutate(
    burden_class = case_when(
      incidence_rate >= quantile(incidence_rate, 0.67, na.rm = TRUE) ~ "High",
      incidence_rate >= quantile(incidence_rate, 0.33, na.rm = TRUE) ~ "Medium",
      TRUE ~ "Low"
    ),
    burden_class = factor(burden_class, levels = c("High", "Medium", "Low"))
  )

# Simulate 70 health facility points across Africa for demonstration
set.seed(4607)
facility_points <- st_sample(africa_health, size = 70, type = "random") |> 
  st_as_sf() |> 
  mutate(facility_id = paste0("FAC_", row_number()),
         type = "Health Facilities")

decision_map <- tm_shape(africa_health) +
  tm_fill(
    "burden_class", 
    fill.scale = tm_scale(values = c("High"="red","Medium"="yellow","Low"="green")),
    fill.legend = tm_legend(title = "Disease Burden", 
                            position = c("left", "bottom")) 
  ) +
  tm_borders() +
  tm_shape(facility_points) +
  tm_dots(
    fill = "type",
    fill.scale = tm_scale_categorical(values ="darkblue"),
    fill.legend = tm_legend(title = "Facilities", 
                            position = c("left", "bottom")),
    size = 0.4
  ) +
  tm_title("High Burden vs. Facility Coverage") 

decision_map
```

::: callout-note
## How the classification works

- `quantile(incidence_rate, 0.67)` returns the value that splits the countries so that roughly two-thirds are below it and one-third above. `case_when()` then labels each country "High", "Medium", or "Low" against the 0.67 and 0.33 cut-points. With 54 countries this produces three groups of 18.
- Converting `burden_class` to a `factor(levels = c("High","Medium","Low"))` fixes the legend order so "High" appears first.
- `st_sample(..., type = "random")` drops 70 random points inside the polygons (reproducible thanks to `set.seed(4607)`), standing in for real health-facility coordinates.
- The map itself layers the burden fill and the facility dots; red polygons with no nearby blue dots are the priority areas.
:::

**Expected output:** A map where red (high-burden) polygons without nearby blue dots immediately signal priority areas for new facilities.

------------------------------------------------------------------------

## Distance / Accessibility Analysis {#sec-distance}

**Purpose:** Quantify the distance from each administrative unit to the nearest health facility.

```{r distance-analysis}
#| warning: false
#| message: false

# Transform both facility points and africa_health sf object to WGS 84 (EPSG:4326)
# This gives true geodesic distances with st_distance
africa_health_geo <- st_transform(africa_health, crs = 4326)
facility_points_geo <- st_transform(facility_points, crs = 4326)

# Calculate centroid of each polygon (using transformed data)
africa_centroids <- st_centroid(africa_health_geo, of_largest_polygon = TRUE)

# Matrix of geodesic distances (meters) between centroids and facility points
dist_matrix <- st_distance(africa_centroids, facility_points_geo)

# Minimum distance per country (convert to km) and append to original dataset
africa_health <- africa_health |>
  mutate(
    dist_to_facility_km = as.numeric(apply(dist_matrix, 1, min)) / 1000
  )

# Summary statistics
summary(africa_health$dist_to_facility_km)

# Map accessibility
access_map <- tm_shape(africa_health) +
  tm_polygons(
    fill = "dist_to_facility_km", 
    fill.scale = tm_scale(values = "brewer.bu_pu"),
    fill.legend = tm_legend(title = "Distance (km)")
        ) +
  tm_shape(facility_points) +
  tm_dots(
    fill = "type",
    fill.scale = tm_scale_categorical(values ="red"),
    fill.legend = tm_legend(title = "Facility"),
    size = 0.4
  ) +
  tm_title("Distance to Nearest Health Facility")

access_map
```

::: callout-note
## How the distance pipeline works

- `st_centroid(..., of_largest_polygon = TRUE)` places one point at the centre of each country (using the largest piece for multi-part countries such as island nations).
- `st_distance(africa_centroids, facility_points_geo)` returns a **matrix** with one row per country and one column per facility — cell `(i, j)` is the distance from country *i* to facility *j*. Here that is a 54 × 70 matrix.
- `apply(dist_matrix, 1, min)` takes the minimum along each *row*, i.e. the distance from each country to its *nearest* facility. Dividing by 1000 converts metres to kilometres.
- The map then shades countries by that nearest-facility distance — the classic accessibility map used to plan service placement.
:::

::: callout-warning
## CRS and Geodesic Distances

When calculating distances, it is crucial to know your CRS.
Because our data was transformed to EPSG:4326 (WGS 84), the `sf` package uses spherical maths to calculate accurate, real-world geodesic distances, returning the results in metres.
Dividing by 1000 gives us kilometres.
:::

::: callout-note
## Projected CRS for local analysis

While the spherical engine handles distance perfectly, if you are doing highly localized spatial modelling or calculating exact polygon areas in Ghana, it is still best practice to reproject your data to a metric coordinate system like UTM Zone 30N (EPSG:32630).
:::

**Expected output:** - Summary statistics of distance (here the median is around 346 km, ranging from roughly 29 km to 2,380 km).
- A map showing increasing distance as darker purple fills.

------------------------------------------------------------------------

## Result Interpretation and Report Generation {#sec-report}

**Purpose:** Produce a reproducible, shareable document integrating code, results, and narrative.

A **reproducible report** is one where the text, numbers, tables, and figures are all generated from code, so a colleague (or future you) can re-run it and get identical results.
R Markdown (`.Rmd`) and Quarto (`.qmd`) are the gold standards for reproducible research.
Below is a minimal Quarto document skeleton you can adapt:

```` markdown
---
title: "Spatial Epidemiology Report: Pancreatic Cancer in Africa"
format: html
editor: visual
---

## Introduction
This report analyzes the spatial distribution ...

## Data
```r
#| echo: false
library(mlspatial)
data(africa_shp)
data(panc_incidence)
```

## Results
The Global Moran's I was 0.074 (inserted via inline code) ...

## Conclusion
High-burden hotspots are concentrated in ...
````

::: callout-note
## The three ingredients of a reproducible report

- **Code chunks** (```` ```{r} ````) run when the document is rendered, so every figure and table is regenerated from the current data.
- **Inline code** embeds an R expression directly in a sentence (wrapped in backticks and prefixed with the letter `r`), so computed numbers update automatically when the data changes instead of being typed by hand.
- **Parameterised reports** use a `params:` block in the YAML header to produce, say, one country report per country from a single template.
:::

**Key features:** - **Code chunks** (```` ```{r} ````) execute during rendering.
- **Inline code** embeds R expressions in the text, updating numbers automatically.
- **Parameterised reports**: Use YAML `params:` to generate country-specific reports from one template.

**Expected output:** An HTML or PDF file containing embedded maps, tables, and narrative — fully reproducible by any colleague who has the `.qmd` file and data.

------------------------------------------------------------------------

## Ethical Considerations {#sec-ethics}

**Purpose:** Reflect on methodological biases and data privacy before disseminating findings.

Spatial epidemiology maps *people*, not just polygons.
Before publishing, it is worth pausing to consider how a map can help — or harm — the communities it describes.

### Privacy Protection

- **Aggregated data**: The `mlspatial` built-in data are already aggregated to country level, protecting individual identity. Never map exact household locations or small-area counts (\< 5 cases) without differential privacy or aggregation.
- **Re-identification risk**: Combining spatial coordinates with demographic details (age, sex, date of diagnosis) can re-identify patients in small villages.

### Modifiable Areal Unit Problem (MAUP)

- **Definition**: Results depend on the boundaries chosen. Country-level analysis may hide sub-national hotspots; conversely, fine-scale analysis may produce unstable rates due to small populations.
- **Mitigation**: Conduct sensitivity analyses at multiple scales (district, region, country) and report how conclusions change.

### Ecological Fallacy

- **Definition**: Relationships observed at the aggregate level do not necessarily hold for individuals. For example, a country with high average temperature and high cancer incidence does not prove that heat causes cancer in individual patients.
- **Mitigation**: Use language like "areas with higher temperature are associated with higher incidence" rather than "temperature causes cancer."

### Spatial Inequality & Resource Allocation

- Maps are political. Highlighting "high burden" regions can attract funding, but it can also stigmatize communities. Always:
  1.  Share results with local health officials before publication.
  2.  Provide confidence intervals or classification uncertainty.
  3.  Pair burden maps with resource maps (as in @sec-overlay-decision) to focus on actionable gaps.

------------------------------------------------------------------------

## Practical Exercise {#sec-m6-exercise}

Apply the module's two core skills, then check your answers against the solution.

1.  Classify `incidence_rate` into **quartiles** (4 groups labelled "Q1" … "Q4") instead of tertiles, and report how many countries fall in each group.
2.  Recompute the nearest-facility distance and report the **median** and **maximum** distance in kilometres.
3.  *(Challenge)* Which country has the largest distance to a facility? Use `which.max()` on the distance column and print that country's name.

::: {.callout-note collapse="true"}
## Model solution

``` r
library(sf)
library(dplyr)
library(mlspatial)

data(africa_shp, package = "mlspatial")
data(panc_incidence, package = "mlspatial")
africa_health <- join_data(
  africa_shp |> mutate(NAME = gsub("Côte d'Ivoire", "Cote d'Ivoire", NAME)),
  panc_incidence, by = "NAME"
)
africa_health <- st_make_valid(africa_health)

set.seed(898)
n <- nrow(africa_health)
africa_health <- africa_health |>
  mutate(incidence_rate = (incidence / TOTPOP) * 100000)

# 1. Quartile classification
africa_health <- africa_health |>
  mutate(burden_q = case_when(
    incidence_rate >= quantile(incidence_rate, 0.75, na.rm = TRUE) ~ "Q4",
    incidence_rate >= quantile(incidence_rate, 0.50, na.rm = TRUE) ~ "Q3",
    incidence_rate >= quantile(incidence_rate, 0.25, na.rm = TRUE) ~ "Q2",
    TRUE ~ "Q1"
  ))
table(africa_health$burden_q)

# 2. Nearest-facility distance
set.seed(4607)
facility_points <- st_sample(africa_health, size = 70, type = "random") |>
  st_as_sf() |>
  mutate(facility_id = paste0("FAC_", row_number()))

centroids <- st_centroid(st_transform(africa_health, 4326),
                         of_largest_polygon = TRUE)
d <- st_distance(centroids, st_transform(facility_points, 4326))
africa_health$dist_km <- as.numeric(apply(d, 1, min)) / 1000

summary(africa_health$dist_km)          # median and max

# 3. Country farthest from a facility
africa_health$NAME[which.max(africa_health$dist_km)]
```
:::

------------------------------------------------------------------------

## Closing Remarks {#sec-closing .unnumbered}

You have now completed a full spatial epidemiology workflow in **R only**:

1.  ✅ Imported and joined spatial and health data
2.  ✅ Created static, interactive, and multi-layer thematic maps
3.  ✅ Detected global and local spatial autocorrelation
4.  ✅ Built OLS and Geographically Weighted Regression models
5.  ✅ Trained and evaluated Random Forest, XGBoost, and SVR models via `mlspatial`
6.  ✅ Integrated disease burden with service availability and generated reproducible reports

**Next steps for the Ghana R User Community:** - Replace the built-in Africa data with Ghana Health Service district-level data.
- Download real WorldClim rasters via `geodata::worldclim_global()`.
- Explore `mlspatial` vignettes: `vignette("mlspatial")` after installing the package.

**Reference:** Azeez, A., & Noel, C.
(2025).
*Predictive Modelling and Spatial Distribution of Pancreatic Cancer in Africa Using Machine Learning-Based Spatial Model*.
Zenodo.
https://doi.org/10.5281/zenodo.16529986
